---
title: "Step 7: MaxEnt Parameter Optimization with ENMeval"
author: "Alexander W. Gofton"
date: today
date-format: "D MMMM YYYY"
format:
  html:
    page-layout: full
    toc: true
    toc-depth: 3
    code-fold: false
    code-tools: true
    embed-resources: true
    theme:
      light: cosmo
      dark: darkly
    fig-width: 12
    fig-height: 8
execute:
  warning: false
  message: false
---

## Overview

This script uses **ENMeval 2.0** to systematically optimize MaxEnt parameters by testing 30 different model configurations:

- **5 feature class combinations**: L, LQ, LQH, LQHP, LQHPT
- **6 regularization multipliers**: β = 0.5, 1.0, 1.5, 2.0, 3.0, 4.0
- **Spatial block cross-validation**: k=5 folds to account for spatial autocorrelation

The best model (lowest AICc) will be used for future climate projections.

**Context from Steps 5-6**:
- Baseline model: AUC = 0.976 (excellent)
- 92% extrapolation in Australia (acceptable, see Step 6b)
- 614 global occurrence points (61% Asia, 10% Australia, 29% other)

## Setup

```{r setup}

# Set Java heap memory to 8GB (adjust based on available RAM)
options(java.parameters = "-Xmx12g")

library(tidyverse)
library(terra)
library(predicts)  # Modern MaxEnt
library(ENMeval)   # Model optimization (v2.0+)
library(rJava)
library(sf)
library(ggplot2)
library(patchwork)
library(beepr)

# Define directories
processed_data_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/processed_data/"
figures_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/figures/"
outputs_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/outputs/"
```

## Load Data

```{r load_data}
# Load occurrence data
occurrences <- readRDS(paste0(processed_data_dir, "occurrences_selected_vars.rds"))

# Load climate data (global extent for training)
bioclim_selected <- readRDS(paste0(processed_data_dir, "bioclim_selected.rds"))

# Load background points (from Step 5)
background_coords <- read.csv(paste0(processed_data_dir, "background_points.csv"))

cat("Data loaded successfully\n")
cat("Occurrence points:", nrow(occurrences), "\n")
cat("Background points:", nrow(background_coords), "\n")
cat("Climate variables:", names(bioclim_selected), "\n")
```

## Prepare Data for ENMeval

```{r prepare_data}
# Extract presence coordinates and remove NAs
presence_coords <- occurrences %>%
  select(lon, lat) %>%
  filter(!is.na(lon), !is.na(lat)) %>%
  as.data.frame()

# Also filter background coordinates for NAs
background_coords_clean <- background_coords %>%
  filter(!is.na(lon), !is.na(lat))

# Convert to sf objects (ENMeval 2.0 works with sf)
presence_sf <- st_as_sf(presence_coords,
                        coords = c("lon", "lat"),
                        crs = 4326)

background_sf <- st_as_sf(background_coords_clean,
                          coords = c("lon", "lat"),
                          crs = 4326)

cat("\nData prepared for ENMeval:\n")
cat("Presence points (sf):", nrow(presence_sf), "\n")
cat("Background points (sf):", nrow(background_sf), "\n")
cat("Points removed due to missing coords:",
    nrow(occurrences) - nrow(presence_sf), "\n")
```

## Configure ENMeval Parameters

```{r enmeval_config}
# Define feature class combinations to test
feature_classes <- c("L", "LQ", "LQH", "LQHP", "LQHPT")

# Define regularization multipliers to test
reg_multipliers <- c(0.5, 1.0, 1.5, 2.0, 3.0, 4.0)

cat("=== ENMEVAL CONFIGURATION ===\n\n")
cat("Feature classes to test:\n")
cat(" L    = Linear only\n")
cat(" LQ   = Linear + Quadratic\n")
cat(" LQH  = Linear + Quadratic + Hinge\n")
cat(" LQHP = Linear + Quadratic + Hinge + Product\n")
cat(" LQHPT = All features (L + Q + H + P + Threshold)\n\n")

cat("Regularization multipliers (β):", paste(reg_multipliers, collapse = ", "), "\n")
cat("Total models to test:", length(feature_classes) * length(reg_multipliers), "\n\n")

cat("Cross-validation strategy: Spatial block (checkerboard2)\n")
cat("Number of folds: 4\n")
cat("Model selection criterion: AICc (lowest = best)\n\n")
```

## Run ENMeval

**Note**: This will take 30-90 minutes depending on your computer.

```{r run_enmeval}
#Starting ENMeval optimization
#This may take 30-90 minutes with 614 global occurrence points

# Check if ENMeval results already generated
enmeval_results_file <- paste0(outputs_dir, "enmeval_results.rds")

if (file.exists(enmeval_results_file)) {
  cat("Loading existing ENMeval results from file...\n")
  enmeval_results <- readRDS(enmeval_results_file)
} else {
  cat("Starting ENMeval optimization...\n")

  # Run ENMeval
  # Set seed for reproducibility
  set.seed(7990)

  # ENMeval with terra::SpatRaster expects regular data frames, not sf objects
  enmeval_results <- ENMevaluate(
    occs = presence_coords,      # Use data frame, not sf
    envs = bioclim_selected,
    bg = background_coords_clean, # Use data frame, not sf
    algorithm = "maxent.jar",
    partitions = "block",
    tune.args = list(
      fc = feature_classes,
      rm = reg_multipliers
    ),
    parallel = TRUE,
    numCores = 4,
    quiet = FALSE
  )

  cat("\n✓ ENMeval optimization complete!\n")
  beep(sound = 8)

  # Save full ENMeval object
  saveRDS(enmeval_results, enmeval_results_file))

  cat("ENMeval results saved to:", enmeval_results_file,  "\n")
}
```


## Extract Results Table

```{r results_table}
# Get results table
results_table <- eval.results(enmeval_results)

# View all results
cat("\n=== ALL MODEL RESULTS ===\n\n")
print(results_table)

# Save results table
write.csv(results_table,
          paste0(outputs_dir, "enmeval_results_table.csv"),
          row.names = FALSE)

cat("\nResults table saved to:", paste0(outputs_dir, "enmeval_results_table.csv"), "\n")
```

## Identify Best Model

```{r best_model}
# Select model with lowest AICc
best_model_idx <- which.min(results_table$AICc)
best_model_row <- results_table[best_model_idx, ]

cat("\n=== BEST MODEL (Lowest AICc) ===\n\n")
cat("Feature Class:", best_model_row$fc, "\n")
cat("Regularization Multiplier:", best_model_row$rm, "\n")
cat("AICc:", round(best_model_row$AICc, 2), "\n")
cat("Delta AICc:", round(best_model_row$delta.AICc, 2), "\n")
cat("AUC (validation):", round(best_model_row$auc.val.avg, 4), "\n")
cat("AUC (training):", round(best_model_row$auc.train, 4), "\n")
cat("AUC difference:", round(best_model_row$auc.train - best_model_row$auc.val.avg, 4), "\n")
cat("Omission rate (validation):", round(best_model_row$or.10p.avg, 4), "\n")

# Check for overfitting
if ((best_model_row$auc.train - best_model_row$auc.val.avg) > 0.1) {
  cat("\n⚠ WARNING: Possible overfitting (AUC diff > 0.1)\n")
} else {
  cat("\n✓ No severe overfitting detected\n")
}

# Compare to baseline
cat("\n=== COMPARISON TO BASELINE ===\n")
cat("Baseline model: fc = LQHPT, rm = 1.0\n")
baseline_idx <- which(results_table$fc == "LQHPT" & results_table$rm == 1.0)

if (length(baseline_idx) > 0) {
  baseline_row <- results_table[baseline_idx, ]
  cat("Baseline AICc:", round(baseline_row$AICc, 2), "\n")
  cat("Baseline AUC (val):", round(baseline_row$auc.val.avg, 4), "\n")
  cat("\nImprovement:\n")
  cat("  Delta AICc:", round(baseline_row$AICc - best_model_row$AICc, 2), "(lower is better)\n")
  cat("  Delta AUC:", round(best_model_row$auc.val.avg - baseline_row$auc.val.avg, 4), "\n")
}
```

## Model Selection Plots

### AICc vs Parameters

```{r aicc_plots}
# AICc by feature class
p1 <- ggplot(results_table, aes(x = fc, y = AICc)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_point(alpha = 0.5, size = 2) +
  geom_hline(yintercept = best_model_row$AICc,
             linetype = "dashed",
             color = "red",
             linewidth = 1) +
  annotate("text",
           x = 1,
           y = best_model_row$AICc,
           label = "Best model",
           vjust = -0.5,
           color = "red",
           fontface = "bold") +
  labs(
    title = "AICc by Feature Class",
    subtitle = "Lower AICc = Better model",
    x = "Feature Class",
    y = "AICc"
  ) +
  theme_minimal(base_size = 12)

# AICc by regularization multiplier
p2 <- ggplot(results_table, aes(x = factor(rm), y = AICc)) +
  geom_boxplot(fill = "coral", alpha = 0.7) +
  geom_point(alpha = 0.5, size = 2) +
  geom_hline(yintercept = best_model_row$AICc,
             linetype = "dashed",
             color = "red",
             linewidth = 1) +
  labs(
    title = "AICc by Regularization Multiplier",
    subtitle = "Higher β = Simpler model",
    x = "Regularization Multiplier (β)",
    y = "AICc"
  ) +
  theme_minimal(base_size = 12)

# Combine plots
combined <- p1 + p2 +
  plot_annotation(
    title = "Model Selection: AICc by Parameters",
    theme = theme(plot.title = element_text(size = 14, face = "bold"))
  )

combined

ggsave(paste0(figures_dir, "07_aicc_by_parameters.png"),
       combined, width = 14, height = 6, dpi = 300)

cat("\nAICc plots saved\n")
```

### AUC vs Complexity

```{r auc_plots}
# Add complexity metric (higher fc + lower rm = more complex)
results_table_complex <- results_table %>%
  mutate(
    fc_num = case_when(
      fc == "L" ~ 1,
      fc == "LQ" ~ 2,
      fc == "LQH" ~ 3,
      fc == "LQHP" ~ 4,
      fc == "LQHPT" ~ 5
    ),
    complexity = fc_num / rm  # Higher = more complex
  )

# AUC validation vs complexity
ggplot(results_table_complex, aes(x = complexity, y = auc.val.avg)) +
  geom_point(aes(color = fc), size = 3, alpha = 0.7) +
  geom_point(data = results_table_complex[best_model_idx, ],
             color = "red",
             size = 6,
             shape = 1,
             stroke = 2) +
  geom_smooth(method = "loess", se = TRUE, color = "gray40", linetype = "dashed") +
  labs(
    title = "Model Performance vs Complexity",
    subtitle = "Red circle = Best model (lowest AICc)",
    x = "Model Complexity (Feature Class / Regularization)",
    y = "AUC (Validation Average)",
    color = "Feature\nClass"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

ggsave(paste0(figures_dir, "07_auc_vs_complexity.png"),
       width = 10, height = 6, dpi = 300)

cat("AUC vs complexity plot saved\n")
```

### Overfitting Analysis

```{r overfitting_plot}
# Calculate AUC difference (overfitting indicator)
results_table_overfit <- results_table %>%
  mutate(auc_diff = auc.train - auc.val.avg)

# Plot overfitting by parameters
ggplot(results_table_overfit, aes(x = fc, y = auc_diff, fill = factor(rm))) +
  geom_col(position = "dodge", alpha = 0.8) +
  geom_hline(yintercept = 0.1,
             linetype = "dashed",
             color = "red",
             linewidth = 1) +
  annotate("text",
           x = 2.5,
           y = 0.1,
           label = "Overfitting threshold (0.1)",
           vjust = -0.5,
           color = "red") +
  scale_fill_viridis_d(name = "Reg. Mult. (β)") +
  labs(
    title = "Overfitting Analysis: Training vs Validation AUC Difference",
    subtitle = "Values > 0.1 indicate possible overfitting",
    x = "Feature Class",
    y = "AUC Difference (Train - Validation)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(paste0(figures_dir, "07_overfitting_analysis.png"),
       width = 12, height = 7, dpi = 300)

cat("Overfitting analysis plot saved\n")
```

## Generate Predictions with Best Model

```{r best_model_prediction}
cat("\nGenerating predictions with best model...\n")

# Get the best model object from ENMeval results
best_model <- eval.models(enmeval_results)[[best_model_idx]]

# Predict to Australia/NZ extent
aus_nz_extent <- ext(110, 180, -48, -10)
bioclim_aus <- crop(bioclim_selected, aus_nz_extent)

# Generate prediction
prediction_optimized <- predict(
  best_model,
  bioclim_aus,
  args = c('outputformat=logistic')
)

cat("Prediction generated\n")
cat("Value range:", round(minmax(prediction_optimized)[1], 3), "to",
    round(minmax(prediction_optimized)[2], 3), "\n")

# Save prediction
writeRaster(
  prediction_optimized,
  paste0(processed_data_dir, "maxent_optimized_prediction_aus.tif"),
  overwrite = TRUE
)

cat("Optimized prediction saved\n")
```

## Visualize Optimized Prediction

```{r visualize_prediction}
library(tidyterra)

# Load occurrence points for overlay
presence_coords <- occurrences %>%
  select(lon, lat) %>%
  filter(lon >= 110, lon <= 180, lat >= -48, lat <= -10)

# Plot
ggplot() +
  geom_spatraster(data = prediction_optimized) +
  geom_point(data = presence_coords,
             aes(x = lon, y = lat),
             color = "black",
             size = 1,
             alpha = 0.7) +
  scale_fill_whitebox_c(
    palette = "muted",
    na.value = "transparent",
    labels = scales::percent,
    name = "Habitat\nSuitability"
  ) +
  coord_sf(xlim = c(110, 180), ylim = c(-48, -10)) +
  labs(
    title = "Optimized MaxEnt Prediction",
    subtitle = paste0("Best model: fc=", best_model_row$fc,
                     ", rm=", best_model_row$rm,
                     " (AUC=", round(best_model_row$auc.val.avg, 3), ")"),
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", size = 14)
  )

ggsave(paste0(figures_dir, "07_optimized_prediction_australia.png"),
       width = 12, height = 8, dpi = 300)

cat("\nOptimized prediction map saved\n")
```


### Leaflet map of optimized model

```{r}
library(leaflet)
library(leaflet.extras2)

# Create color palette (white to red)
pal <- colorNumeric(
  palette = "Reds",
  domain = values(prediction_optimized, na.rm = TRUE),
  na.color = "transparent"
)

# Create leaflet map
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(
    prediction_optimized,
    colors = pal,
    opacity = 0.9,
    group = "Habitat Suitability"
  ) |>
  addCircleMarkers(
    data = presence_coords,
    lng = ~lon,
    lat = ~lat,
    radius = 4,
    color = "black",
    fillColor = "black",
    fillOpacity = 0.7,
    stroke = TRUE,
    weight = 1,
    group = "Occurrence Points",
    popup = ~paste("Lon:", round(lon, 3), "<br>Lat:", round(lat, 3))
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal,
    values = values(prediction_optimized, na.rm = TRUE),
    title = "Habitat<br>Suitability",
    labFormat = labelFormat(
      suffix = "%",
      transform = function(x) round(x * 100)
    )
  ) |>
  addLayersControl(
    overlayGroups = c("Habitat Suitability", "Occurrence Points"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  setView(lng = 145, lat = -25, zoom = 4)

# Save as HTML
library(htmlwidgets)
saveWidget(
  leaflet() |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addRasterImage(prediction_optimized, colors = pal, opacity = 0.7,
                   group = "Habitat Suitability") |>
    addCircleMarkers(data = presence_coords, lng = ~lon, lat = ~lat,
                     radius = 4, color = "black", fillColor = "black",
                     fillOpacity = 0.7, stroke = TRUE, weight = 1,
                     group = "Occurrence Points",
                     popup = ~paste("Lon:", round(lon, 3), "<br>Lat:", round(lat, 3))) |>
    addLegend(position = "bottomright", pal = pal,
              values = values(prediction_optimized, na.rm = TRUE),
              title = "Habitat<br>Suitability",
              labFormat = labelFormat(suffix = "%",
                                     transform = function(x) round(x * 100))) |>
    addLayersControl(overlayGroups = c("Habitat Suitability", "Occurrence Points"),
                     options = layersControlOptions(collapsed = FALSE)) |>
    setView(lng = 145, lat = -25, zoom = 4),
  file = paste0(figures_dir, "07_interactive_optimized_prediction.html"),
  selfcontained = TRUE
)

cat("\nInteractive map saved to:",
    paste0(figures_dir, "07_interactive_optimized_prediction.html"), "\n")
```

## Compare Baseline vs Optimized

```{r comparison}
# Load baseline prediction from Step 5
prediction_baseline <- rast(paste0(processed_data_dir, "maxent_baseline_prediction_aus.tif"))

# Calculate difference
prediction_diff <- prediction_optimized - prediction_baseline

# Summary statistics
cat("\n=== PREDICTION COMPARISON ===\n\n")
cat("Baseline (fc=LQHPT, rm=1.0):\n")
cat("  Mean suitability:", round(mean(values(prediction_baseline, na.rm = TRUE)), 3), "\n")
cat("  SD:", round(sd(values(prediction_baseline, na.rm = TRUE)), 3), "\n\n")

cat("Optimized (fc=", best_model_row$fc, ", rm=", best_model_row$rm, "):\n", sep = "")
cat("  Mean suitability:", round(mean(values(prediction_optimized, na.rm = TRUE)), 3), "\n")
cat("  SD:", round(sd(values(prediction_optimized, na.rm = TRUE)), 3), "\n\n")

cat("Difference (Optimized - Baseline):\n")
cat("  Mean difference:", round(mean(values(prediction_diff, na.rm = TRUE)), 3), "\n")
cat("  Max increase:", round(max(values(prediction_diff, na.rm = TRUE)), 3), "\n")
cat("  Max decrease:", round(min(values(prediction_diff, na.rm = TRUE)), 3), "\n")

# Visualize difference
ggplot() +
  geom_spatraster(data = prediction_diff) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "transparent",
    name = "Suitability\nChange",
    limits = c(-0.3, 0.3)
  ) +
  coord_sf(xlim = c(110, 180), ylim = c(-48, -10)) +
  labs(
    title = "Prediction Change: Optimized vs Baseline",
    subtitle = "Red = Higher suitability | Blue = Lower suitability",
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

ggsave(paste0(figures_dir, "07_prediction_difference.png"),
       width = 12, height = 8, dpi = 300)

cat("\nDifference map saved\n")
```

```{r leaflet_model_comparison}
library(leaflet)
library(htmlwidgets)

# Create diverging color palette (blue-white-red)
pal_diff <- colorNumeric(
  palette = colorRampPalette(c("blue", "white", "red"))(100),
  domain = c(-0.3, 0.3),
  na.color = "transparent"
)

# Create leaflet map
leaflet() |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addRasterImage(
    prediction_diff,
    colors = pal_diff,
    opacity = 0.8,
    group = "Suitability Change"
  ) |>
  addCircleMarkers(
    data = presence_coords,
    lng = ~lon,
    lat = ~lat,
    radius = 2,
    color = "black",
    fillColor = "black",
    fillOpacity = 0.6,
    stroke = TRUE,
    weight = 1,
    group = "Occurrence Points",
    popup = ~paste("Lon:", round(lon, 3), "<br>Lat:", round(lat, 3))
  ) |>
  addLegend(
    position = "bottomright",
    pal = pal_diff,
    values = c(-0.3, 0.3),
    title = "Suitability<br>Change",
    labFormat = labelFormat(
      prefix = "",
      suffix = "",
      transform = function(x) round(x, 2)
    )
  ) |>
  addLayersControl(
    overlayGroups = c("Suitability Change", "Occurrence Points"),
    options = layersControlOptions(collapsed = FALSE)
  ) |>
  setView(lng = 145, lat = -25, zoom = 4)

# Save as HTML
saveWidget(
  leaflet() |>
    addProviderTiles(providers$CartoDB.Positron) |>
    addRasterImage(prediction_diff, colors = pal_diff, opacity = 0.8,
                   group = "Suitability Change") |>
    addCircleMarkers(data = presence_coords, lng = ~lon, lat = ~lat,
                     radius = 3, color = "black", fillColor = "yellow",
                     fillOpacity = 0.6, stroke = TRUE, weight = 1,
                     group = "Occurrence Points",
                     popup = ~paste("Lon:", round(lon, 3), "<br>Lat:", round(lat, 3))) |>
    addLegend(position = "bottomright", pal = pal_diff,
              values = c(-0.3, 0.3),
              title = "Suitability<br>Change",
              labFormat = labelFormat(prefix = "", suffix = "",
                                     transform = function(x) round(x, 2))) |>
    addLayersControl(overlayGroups = c("Suitability Change", "Occurrence Points"),
                     options = layersControlOptions(collapsed = FALSE)) |>
    setView(lng = 145, lat = -25, zoom = 4),
  file = paste0(figures_dir, "07_interactive_prediction_difference.html"),
  selfcontained = TRUE
)

cat("\nInteractive difference map saved to:",
    paste0(figures_dir, "07_interactive_prediction_difference.html"), "\n")

```

## Save Best Model

```{r save_best_model}
# Save the best model object
saveRDS(best_model, paste0(outputs_dir, "maxent_optimized_model.rds"))

# Save model metadata
model_metadata <- data.frame(
  feature_class = best_model_row$fc,
  regularization_multiplier = best_model_row$rm,
  aicc = best_model_row$AICc,
  delta_aicc = best_model_row$delta.AICc,
  auc_validation = best_model_row$auc.val.avg,
  auc_training = best_model_row$auc.train,
  auc_difference = best_model_row$auc.train - best_model_row$auc.val.avg,
  omission_rate = best_model_row$or.10p.avg,
  n_parameters = best_model_row$ncoef
)

write.csv(model_metadata,
          paste0(outputs_dir, "optimized_model_metadata.csv"),
          row.names = FALSE)

cat("\n=== OUTPUTS SAVED ===\n\n")
cat("✓ Best model object:", paste0(outputs_dir, "maxent_optimized_model.rds"), "\n")
cat("✓ Model metadata:", paste0(outputs_dir, "optimized_model_metadata.csv"), "\n")
cat("✓ ENMeval results:", paste0(outputs_dir, "enmeval_results.rds"), "\n")
cat("✓ Results table:", paste0(outputs_dir, "enmeval_results_table.csv"), "\n")
cat("✓ Optimized prediction:", paste0(processed_data_dir, "maxent_optimized_prediction_aus.tif"), "\n")
cat("✓ Figures:", paste0(figures_dir, "07_*.png"), "\n")
```

## Summary and Next Steps

```{r summary}
cat("\n╔════════════════════════════════════════════════════════════════╗\n")
cat("║     STEP 7 COMPLETE - MAXENT OPTIMIZATION                     ║\n")
cat("╚════════════════════════════════════════════════════════════════╝\n\n")

cat("=== OPTIMIZATION SUMMARY ===\n\n")
cat("Models tested:", nrow(results_table), "\n")
cat("  - Feature classes:", length(feature_classes), "\n")
cat("  - Regularization multipliers:", length(reg_multipliers), "\n")
cat("  - Cross-validation folds: 4 (spatial block)\n\n")

cat("=== BEST MODEL SELECTED ===\n\n")
cat("Selection criterion: Lowest AICc\n")
cat("Feature class:", best_model_row$fc, "\n")
cat("Regularization multiplier:", best_model_row$rm, "\n")
cat("AICc:", round(best_model_row$AICc, 2), "\n")
cat("AUC (validation):", round(best_model_row$auc.val.avg, 4), "\n")
cat("AUC (training):", round(best_model_row$auc.train, 4), "\n")
cat("Overfitting (AUC diff):", round(best_model_row$auc.train - best_model_row$auc.val.avg, 4), "\n")
cat("Omission rate:", round(best_model_row$or.10p.avg, 4), "\n\n")

cat("=== KEY FINDINGS ===\n\n")

# Interpret feature class
if (best_model_row$fc == "L") {
  cat("1. Model selected LINEAR features only\n")
  cat("   → Simplest model, linear relationships between climate and suitability\n")
} else if (best_model_row$fc == "LQ") {
  cat("1. Model selected LINEAR + QUADRATIC features\n")
  cat("   → Allows for unimodal (peaked) responses to climate\n")
} else if (best_model_row$fc == "LQH") {
  cat("1. Model selected LINEAR + QUADRATIC + HINGE features\n")
  cat("   → Moderate complexity, allows for threshold responses\n")
} else {
  cat("1. Model selected complex feature set:", best_model_row$fc, "\n")
  cat("   → High complexity model with interactions\n")
}

# Interpret regularization
if (best_model_row$rm < 1.0) {
  cat("\n2. Low regularization selected (β =", best_model_row$rm, ")\n")
  cat("   → Model prefers more constrained fit to training data\n")
} else if (best_model_row$rm > 2.0) {
  cat("\n2. High regularization selected (β =", best_model_row$rm, ")\n")
  cat("   → Model prefers smoother, more generalized predictions\n")
} else {
  cat("\n2. Moderate regularization selected (β =", best_model_row$rm, ")\n")
  cat("   → Balanced complexity\n")
}

# Performance
cat("\n3. Model performance:\n")
if (best_model_row$auc.val.avg >= 0.9) {
  cat("   → EXCELLENT discrimination (AUC = ", round(best_model_row$auc.val.avg, 3), ")\n", sep = "")
} else if (best_model_row$auc.val.avg >= 0.8) {
  cat("   → GOOD discrimination (AUC = ", round(best_model_row$auc.val.avg, 3), ")\n", sep = "")
} else {
  cat("   → ACCEPTABLE discrimination (AUC = ", round(best_model_row$auc.val.avg, 3), ")\n", sep = "")
}

if ((best_model_row$auc.train - best_model_row$auc.val.avg) < 0.05) {
  cat("   → Minimal overfitting detected\n")
} else if ((best_model_row$auc.train - best_model_row$auc.val.avg) < 0.1) {
  cat("   → Acceptable overfitting levels\n")
} else {
  cat("   → ⚠ Some overfitting detected (consider simpler model)\n")
}

cat("\n=== NEXT STEPS ===\n\n")
cat("→ Step 8: Final model validation with regional cross-validation\n")
cat("→ Step 9: Download future climate data (3 GCMs × 2 SSPs × 2 periods)\n")
cat("→ Step 10: Project optimized model to 12 future scenarios\n")
cat("→ Step 11: Create publication-quality maps and metrics\n\n")

cat("═══════════════════════════════════════════════════════════════\n")

beep(sound = 10)
```

## Session Information

```{r session_info}
sessionInfo()
```
