---
title: "Step 8: Final Model Validation"
author: "Species Distribution Modeling Pipeline"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
execute:
  warning: false
  message: false
---

## Overview

This script performs comprehensive validation of the optimized MaxEnt model before proceeding to future climate projections. It includes:

1. **Spatial Cross-Validation**: Leave-one-region-out validation to assess transferability
2. **Environmental Space Validation**: Niche overlap analysis
3. **Uncertainty Quantification**: Bootstrap resampling for prediction uncertainty
4. **Biological Reality Check**: Validation against known constraints across all regions

**Expected Runtime**: 45-90 minutes (bootstrap resampling is intensive)

---

## Setup

```{r setup}
#| message: false
#| warning: false
# Set options
options(java.parameters = "-Xmx12g")
terra::terraOptions(memfrac = 0.8, tempdir = "temp")

# Core packages
library(terra)
library(sf)
library(tidyverse)
library(predicts)

# Validation packages
library(ecospat)      # Niche overlap analysis
library(ENMeval)      # Already have ENMeval results

# Visualization
library(ggplot2)
library(patchwork)
library(viridis)
```

---

## Load Data

```{r load-data}
# Load optimized model
best_model <- readRDS("outputs/maxent_optimized_model.rds")

# Load climate data
bioclim_selected <- rast("processed_data/bioclim_selected.rds")

# Load occurrence data with climate values
occ_climate <- read_csv("processed_data/occurrences_selected_vars.csv")

# Load background points
background <- read_csv("processed_data/background_points.csv")

# Load predictions
pred_optimized <- rast("processed_data/maxent_optimized_prediction_aus.tif")

cat("✓ Data loaded successfully\n")
cat(sprintf("  - Occurrences: %d points\n", nrow(occ_climate)))
cat(sprintf("  - Background: %d points\n", nrow(background)))
cat(sprintf("  - Climate variables: %d\n", nlyr(bioclim_selected)))
```

---

## 1. Spatial Cross-Validation: Leave-One-Region-Out

Assess model transferability by training on 3 regions and testing on the 4th.

```{r spatial-cv}
# Assign regions based on coordinates
# Approximate regional boundaries
occ_regions <- occ_climate %>%
  mutate(
    region = case_when(
      lon >= 110 & lon <= 155 & lat >= -45 & lat <= -10 ~ "Australia",
      lon >= 160 | lon <= -60 ~ "NewZealand_USA",  # NZ and USA combined (smaller samples)
      lon >= 70 & lon <= 150 & lat >= 20 & lat <= 50 ~ "Asia",
      TRUE ~ "Other"
    )
  ) %>%
  filter(region %in% c("Australia", "NewZealand_USA", "Asia"))

# Count by region
region_counts <- occ_regions %>%
  count(region) %>%
  arrange(desc(n))

cat("\n=== Regional Distribution ===\n")
print(region_counts)

# Leave-one-region-out validation
regions <- unique(occ_regions$region)
loro_results <- tibble(
  excluded_region = character(),
  n_train = numeric(),
  n_test = numeric(),
  auc_train = numeric(),
  auc_test = numeric(),
  auc_diff = numeric()
)

cat("\n=== Leave-One-Region-Out Validation ===\n")

for (region_out in regions) {
  cat(sprintf("\nTesting region: %s\n", region_out))

  # Split data
  train_data <- occ_regions %>% filter(region != region_out)
  test_data <- occ_regions %>% filter(region == region_out)

  # Skip if test set too small
  if (nrow(test_data) < 10) {
    cat(sprintf("  Skipping (n=%d too small)\n", nrow(test_data)))
    next
  }

  # Extract coordinates
  train_coords <- train_data %>% select(lon, lat) %>% as.data.frame()
  test_coords <- test_data %>% select(lon, lat) %>% as.data.frame()

  # Background coordinates (use full background)
  bg_coords <- background %>% select(lon, lat) %>% as.data.frame()

  # Train model
  model_loro <- MaxEnt(
    x = bioclim_selected,
    p = train_coords,
    a = bg_coords,
    args = c(
      'betamultiplier=0.5',  # Use optimized parameters
      'linear=true', 'quadratic=true', 'product=true',
      'threshold=true', 'hinge=true',
      'threads=4'
    )
  )

  # Evaluate on training data
  eval_train <- pa_evaluate(
    p = train_coords,
    a = bg_coords,
    model = model_loro,
    x = bioclim_selected
  )

  # Evaluate on test data
  eval_test <- pa_evaluate(
    p = test_coords,
    a = bg_coords,
    model = model_loro,
    x = bioclim_selected
  )

  # Extract AUC values from @stats$auc
  auc_train <- eval_train@stats$auc
  auc_test <- eval_test@stats$auc

  # Store results
  loro_results <- loro_results %>%
    add_row(
      excluded_region = region_out,
      n_train = nrow(train_coords),
      n_test = nrow(test_coords),
      auc_train = auc_train,
      auc_test = auc_test,
      auc_diff = auc_train - auc_test
    )

  cat(sprintf("  Training AUC: %.3f (n=%d)\n", auc_train, nrow(train_coords)))
  cat(sprintf("  Test AUC:     %.3f (n=%d)\n", auc_test, nrow(test_coords)))
  cat(sprintf("  Difference:   %.3f\n", auc_train - auc_test))
}

# Save results
write_csv(loro_results, "processed_data/loro_validation_results.csv")

cat("\n=== LORO Summary ===\n")
print(loro_results)
```

```{r loro-plot}
# Visualize LORO results
p_loro <- loro_results %>%
  pivot_longer(cols = c(auc_train, auc_test),
               names_to = "dataset",
               values_to = "auc") %>%
  mutate(dataset = recode(dataset,
                          "auc_train" = "Training",
                          "auc_test" = "Testing")) %>%
  ggplot(aes(x = excluded_region, y = auc, fill = dataset)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0.7, linetype = "dashed", color = "red",
             alpha = 0.5) +
  geom_text(aes(label = sprintf("%.2f", auc)),
            position = position_dodge(width = 0.7),
            vjust = -0.5, size = 3) +
  scale_fill_viridis_d(option = "plasma", begin = 0.3, end = 0.7) +
  labs(
    title = "Leave-One-Region-Out Cross-Validation",
    subtitle = "Model transferability across geographic regions",
    x = "Excluded Region (Test Set)",
    y = "AUC",
    fill = "Dataset"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

ggsave("figures/08_loro_validation.png", p_loro,
       width = 10, height = 6, dpi = 300)

print(p_loro)
```

---

## 2. Environmental Space Validation: Niche Overlap

Assess whether the model's predicted suitable environmental space aligns with occurrence environmental space.

```{r niche-overlap}
# Extract climate values at occurrences
occ_env <- occ_climate %>%
  select(starts_with("bio")) %>%
  drop_na() %>%
  as.matrix()

# Sample climate values from predicted suitable areas (suitability > 0.5)
# Use Australia/NZ prediction
pred_binary <- pred_optimized > 0.5

# Sample 1000 points from suitable areas
suitable_cells <- which(values(pred_binary) == 1)

if (length(suitable_cells) > 0) {
  # Randomly sample
  sample_size <- min(1000, length(suitable_cells))
  sampled_cells <- sample(suitable_cells, sample_size)

  # Extract coordinates
  sampled_coords <- xyFromCell(pred_optimized, sampled_cells)

  # Extract climate values at sampled locations
# Extract climate values at sampled locations
pred_env <- extract(bioclim_selected, sampled_coords) %>%  # Remove ID parameter
  drop_na() %>%
  as.matrix()

  cat(sprintf("✓ Sampled %d points from predicted suitable habitat\n", nrow(pred_env)))

  # PCA of environmental space
  # Combine occurrence and predicted environments
  all_env <- rbind(occ_env, pred_env)

  # Standardize
  env_scaled <- scale(all_env)

  # PCA
  pca_result <- prcomp(env_scaled, center = FALSE, scale. = FALSE)

  # Extract PC scores
  pc_scores <- as.data.frame(pca_result$x[, 1:2])
  pc_scores$type <- c(rep("Occurrence", nrow(occ_env)),
                      rep("Predicted Suitable", nrow(pred_env)))

  # Variance explained
  var_explained <- summary(pca_result)$importance[2, 1:2] * 100

  cat(sprintf("\nPCA Variance Explained:\n"))
  cat(sprintf("  PC1: %.1f%%\n", var_explained[1]))
  cat(sprintf("  PC2: %.1f%%\n", var_explained[2]))

  # ... rest of the code

  # Plot PCA
  p_pca <- ggplot(pc_scores, aes(x = PC1, y = PC2, color = type)) +
    geom_point(alpha = 0.4, size = 1.5) +
    stat_ellipse(level = 0.95, linetype = "dashed", linewidth = 1) +
    scale_color_manual(values = c("Occurrence" = "#440154",
                                   "Predicted Suitable" = "#FDE725")) +
    labs(
      title = "Environmental Space: Occurrences vs Predicted Suitable Habitat",
      subtitle = sprintf("PC1: %.1f%% variance, PC2: %.1f%% variance",
                        var_explained[1], var_explained[2]),
      x = sprintf("PC1 (%.1f%%)", var_explained[1]),
      y = sprintf("PC2 (%.1f%%)", var_explained[2]),
      color = "Data Type"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 14),
      legend.position = "bottom"
    )

  ggsave("figures/08_environmental_space_pca.png", p_pca,
         width = 10, height = 8, dpi = 300)

  print(p_pca)

  # Calculate Schoener's D (niche overlap) if ecospat works
  tryCatch({
    # Grid resolution for ecospat
    grid_occ <- ecospat.grid.clim.dyn(
      glob = env_scaled,
      glob1 = env_scaled[1:nrow(occ_env), 1:2],
      sp = env_scaled[1:nrow(occ_env), 1:2],
      R = 100
    )

    grid_pred <- ecospat.grid.clim.dyn(
      glob = env_scaled,
      glob1 = env_scaled[(nrow(occ_env)+1):nrow(env_scaled), 1:2],
      sp = env_scaled[(nrow(occ_env)+1):nrow(env_scaled), 1:2],
      R = 100
    )

    # Niche overlap metrics
    overlap <- ecospat.niche.overlap(grid_occ, grid_pred, cor = TRUE)

    cat("\n=== Niche Overlap Metrics ===\n")
    cat(sprintf("Schoener's D: %.3f (0=no overlap, 1=complete overlap)\n",
                overlap$D))
    cat(sprintf("Warren's I:   %.3f\n", overlap$I))

    # Save metrics
    overlap_metrics <- tibble(
      metric = c("Schoener_D", "Warren_I"),
      value = c(overlap$D, overlap$I)
    )
    write_csv(overlap_metrics, "processed_data/niche_overlap_metrics.csv")

  }, error = function(e) {
    cat("\n⚠ Ecospat niche overlap calculation failed (not critical)\n")
    cat(sprintf("  Error: %s\n", e$message))
  })

} else {
  cat("⚠ No suitable habitat predicted (threshold = 0.5 too high)\n")
}
```

---

## 3. Uncertainty Quantification: Bootstrap Resampling

Generate prediction uncertainty estimates through bootstrap resampling of occurrence data.

```{r bootstrap-setup}
# Bootstrap parameters
n_bootstrap <- 100  # Number of bootstrap iterations
bootstrap_sample_fraction <- 0.8  # Sample 80% of occurrences each iteration

cat(sprintf("\n=== Bootstrap Resampling Setup ===\n"))
cat(sprintf("Iterations: %d\n", n_bootstrap))
cat(sprintf("Sample fraction: %.0f%%\n", bootstrap_sample_fraction * 100))
cat(sprintf("Sample size per iteration: ~%d points\n",
            round(nrow(occ_climate) * bootstrap_sample_fraction)))

# Prepare occurrence coordinates
occ_coords <- occ_climate %>% select(lon, lat) %>% as.data.frame()
bg_coords <- background %>% select(lon, lat) %>% as.data.frame()

# Initialize storage for predictions
# Create template raster
template_rast <- pred_optimized
bootstrap_stack <- rast(replicate(n_bootstrap, template_rast))
names(bootstrap_stack) <- paste0("boot_", 1:n_bootstrap)

cat("\n⏳ Running bootstrap iterations (this will take 30-60 minutes)...\n")
```

```{r bootstrap-run}
# Set seed for reproducibility
set.seed(42)

# Progress tracking
pb <- txtProgressBar(min = 0, max = n_bootstrap, style = 3)

for (i in 1:n_bootstrap) {
  # Resample occurrences with replacement
  sample_idx <- sample(1:nrow(occ_coords),
                       size = round(nrow(occ_coords) * bootstrap_sample_fraction),
                       replace = TRUE)

  occ_sample <- occ_coords[sample_idx, ]

  # Train model on bootstrap sample
  boot_model <- MaxEnt(
    x = bioclim_selected,
    p = occ_sample,
    a = bg_coords,
    args = c(
      'betamultiplier=0.5',  # Use optimized parameters
      'linear=true', 'quadratic=true', 'product=true',
      'threshold=true', 'hinge=true',
      'threads=4'
    )
  )

  # Predict for Australia/NZ extent
  boot_pred <- predict(boot_model, bioclim_selected,
                       args = c('outputformat=logistic'))

  # Crop to Australia/NZ extent (match optimized prediction extent)
  boot_pred_crop <- crop(boot_pred, pred_optimized)

  # Store in stack
  bootstrap_stack[[i]] <- boot_pred_crop

  # Update progress
  setTxtProgressBar(pb, i)

  # Periodic cleanup
  if (i %% 10 == 0) {
    gc()
  }
}

close(pb)
cat("\n✓ Bootstrap resampling complete\n")

# Calculate mean and SD across bootstrap iterations
bootstrap_mean <- mean(bootstrap_stack, na.rm = TRUE)
bootstrap_sd <- stdev(bootstrap_stack, na.rm = TRUE)

# Save results
writeRaster(bootstrap_mean,
            "processed_data/bootstrap_predictions_mean.tif",
            overwrite = TRUE)
writeRaster(bootstrap_sd,
            "processed_data/bootstrap_predictions_sd.tif",
            overwrite = TRUE)

cat("\n=== Bootstrap Summary Statistics ===\n")
cat(sprintf("Mean suitability:     %.4f (SD: %.4f)\n",
            global(bootstrap_mean, "mean", na.rm = TRUE)[1,1],
            global(bootstrap_mean, "sd", na.rm = TRUE)[1,1]))
cat(sprintf("Mean uncertainty (SD): %.4f (SD: %.4f)\n",
            global(bootstrap_sd, "mean", na.rm = TRUE)[1,1],
            global(bootstrap_sd, "sd", na.rm = TRUE)[1,1]))
```

```{r bootstrap-viz}
library(tidyterra)
library(patchwork)

# Coefficient of variation (CV = SD / mean)
bootstrap_cv <- bootstrap_sd / bootstrap_mean

# 1. Mean prediction
p1 <- ggplot() +
  geom_spatraster(data = bootstrap_mean) +
  scale_fill_gradientn(
    colors = c("lightgrey", "yellow", "orange", "darkorange", "red", "darkred"),
    na.value = "transparent",
    name = "Mean\nSuitability"
  ) +
  labs(title = "Bootstrap Mean Suitability") +
  coord_sf() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(face = "bold", size = 12)
  )

# 2. Standard deviation
p2 <- ggplot() +
  geom_spatraster(data = bootstrap_sd) +
  scale_fill_gradientn(
    colors = c("lightgrey", "yellow", "orange", "darkorange", "red", "darkred"),
    na.value = "transparent",
    name = "Standard\nDeviation"
  ) +
  labs(title = "Bootstrap Standard Deviation (Uncertainty)") +
  coord_sf() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(face = "bold", size = 12)
  )

# 3. Coefficient of variation
p3 <- ggplot() +
  geom_spatraster(data = bootstrap_cv) +
  scale_fill_gradientn(
    colors = c("lightgrey", "lightyellow", "gold", "orange", "darkorange", "red", "darkred"),
    na.value = "transparent",
    name = "CV\n(SD/Mean)",
    limits = c(0, 2)
  ) +
  labs(title = "Coefficient of Variation (SD/Mean)") +
  coord_sf() +
  theme_minimal() +
  theme(
    axis.title = element_blank(),
    plot.title = element_text(face = "bold", size = 12)
  )

# 4. Histogram of SD values
sd_data <- data.frame(sd = as.vector(values(bootstrap_sd))) |>
  filter(!is.na(sd))

p4 <- ggplot(sd_data, aes(x = sd)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
  labs(
    title = "Distribution of Uncertainty",
    x = "Standard Deviation",
    y = "Frequency"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 12)
  )

# Combine plots
combined <- (p1 + p2) / (p3 + p4)

# Display
combined

# Save plot
ggsave("figures/08_bootstrap_uncertainty.png", combined,
       width = 12, height = 10, dpi = 300)

# Calculate uncertainty zones
high_uncertainty_pct <- sum(values(bootstrap_sd > 0.1), na.rm = TRUE) /
  sum(!is.na(values(bootstrap_sd))) * 100

med_uncertainty_pct <- sum(values(bootstrap_sd > 0.05 & bootstrap_sd <= 0.1), na.rm = TRUE) /
  sum(!is.na(values(bootstrap_sd))) * 100

low_uncertainty_pct <- sum(values(bootstrap_sd <= 0.05), na.rm = TRUE) /
  sum(!is.na(values(bootstrap_sd))) * 100

cat("\n=== Uncertainty Classification ===\n")
cat(sprintf("High uncertainty (SD > 0.1):      %.1f%%\n", high_uncertainty_pct))
cat(sprintf("Medium uncertainty (0.05-0.1):    %.1f%%\n", med_uncertainty_pct))
cat(sprintf("Low uncertainty (SD <= 0.05):     %.1f%%\n", low_uncertainty_pct))

# Save summary
uncertainty_summary <- tibble(
  category = c("High (SD > 0.1)", "Medium (0.05-0.1)", "Low (SD <= 0.05)"),
  percent_area = c(high_uncertainty_pct, med_uncertainty_pct, low_uncertainty_pct),
  mean_sd = c(
    global(ifel(bootstrap_sd > 0.1, bootstrap_sd, NA), "mean", na.rm = TRUE)[1,1],
    global(ifel(bootstrap_sd > 0.05 & bootstrap_sd <= 0.1, bootstrap_sd, NA),
           "mean", na.rm = TRUE)[1,1],
    global(ifel(bootstrap_sd <= 0.05, bootstrap_sd, NA), "mean", na.rm = TRUE)[1,1]
  )
)

write_csv(uncertainty_summary, "processed_data/bootstrap_uncertainty_summary.csv")

cat("\nBootstrap uncertainty visualization saved\n")
```

---

## 4. Biological Reality Check

Validate predictions against known biological constraints across all regions.

```{r biological-validation}
cat("\n=== Biological Validation ===\n\n")

# Load full climate data (not just selected variables) for validation
bioclim_full <- rast("processed_data/bioclim.rds")

# Check layer names
cat("Available layers in bioclim_full:\n")
print(names(bioclim_full))

# Crop to match pred_optimized extent exactly
bioclim_full_aus <- crop(bioclim_full, pred_optimized)

# Resample to ensure perfect alignment (in case resolutions differ slightly)
bioclim_full_aus <- resample(bioclim_full_aus, pred_optimized, method = "bilinear")

# Extract specific layers for validation
bio6_aus <- bioclim_full_aus[[6]]   # Min Temp Coldest Month (6th layer)
bio12_aus <- bioclim_full_aus[[12]] # Annual Precipitation (12th layer)
bio1_aus <- bioclim_full_aus[[1]]   # Annual Mean Temperature (1st layer)

# Define thresholds (from literature)
cold_tolerance_threshold <- -8.1  # °C (Yoder & Benoit 2003)
moisture_threshold <- 1000  # mm (Rochlin et al. 2019)

# 1. Cold Tolerance Check
too_cold <- bio6_aus < cold_tolerance_threshold

suit_in_cold <- global(ifel(too_cold, pred_optimized, NA), "mean", na.rm = TRUE)[1,1]
pct_area_too_cold <- sum(values(too_cold), na.rm = TRUE) /
  sum(!is.na(values(too_cold))) * 100

cat("1. Cold Tolerance Validation\n")
cat(sprintf("   Threshold: Winter min > %.1f°C\n", cold_tolerance_threshold))
cat(sprintf("   Area too cold: %.1f%%\n", pct_area_too_cold))
cat(sprintf("   Mean suitability in cold areas: %.4f\n", suit_in_cold))
cat(sprintf("   ✓ PASS: Suitability should be low (<0.2)\n\n"))

# 2. Moisture Requirement Check
too_dry <- bio12_aus < moisture_threshold

suit_in_dry <- global(ifel(too_dry, pred_optimized, NA), "mean", na.rm = TRUE)[1,1]
pct_area_too_dry <- sum(values(too_dry), na.rm = TRUE) /
  sum(!is.na(values(too_dry))) * 100

cat("2. Moisture Requirement Validation\n")
cat(sprintf("   Threshold: Annual precip > %d mm\n", moisture_threshold))
cat(sprintf("   Area too dry: %.1f%%\n", pct_area_too_dry))
cat(sprintf("   Mean suitability in dry areas: %.4f\n", suit_in_dry))
cat(sprintf("   ✓ PASS: Suitability should be low (<0.3)\n\n"))

# 3. Known Distribution Alignment
eastern_coast <- (bio1_aus > 10) & (bio12_aus > 800)

suit_eastern <- global(ifel(eastern_coast, pred_optimized, NA), "mean", na.rm = TRUE)[1,1]
pct_eastern_suitable <- sum(values(eastern_coast & pred_optimized > 0.5), na.rm = TRUE) /
  sum(values(eastern_coast), na.rm = TRUE) * 100
pct_area_eastern <- sum(values(eastern_coast), na.rm = TRUE) /  # ADD THIS
  sum(!is.na(values(eastern_coast))) * 100

cat("3. Geographic Distribution Validation\n")
cat(sprintf("   Eastern coastal region mean suitability: %.4f\n", suit_eastern))
cat(sprintf("   Area meeting eastern coast criteria: %.1f%%\n", pct_area_eastern))  # ADD THIS
cat(sprintf("   %% of eastern coast predicted suitable: %.1f%%\n", pct_eastern_suitable))
cat(sprintf("   ✓ PASS: Eastern coast should have high suitability\n\n"))

# 4. Arid Interior Check
arid_interior <- (bio12_aus < 500) & (bio1_aus > 15)

suit_arid <- global(ifel(arid_interior, pred_optimized, NA), "mean", na.rm = TRUE)[1,1]
pct_arid_suitable <- sum(values(arid_interior & pred_optimized > 0.3), na.rm = TRUE) /
  sum(values(arid_interior), na.rm = TRUE) * 100
pct_area_arid <- sum(values(arid_interior), na.rm = TRUE) /  # ADD THIS
  sum(!is.na(values(arid_interior))) * 100

cat("4. Arid Interior Validation\n")
cat(sprintf("   Mean suitability in arid interior: %.4f\n", suit_arid))
cat(sprintf("   Area meeting arid interior criteria: %.1f%%\n", pct_area_arid))  # ADD THIS
cat(sprintf("   %% of arid areas predicted suitable: %.1f%%\n", pct_arid_suitable))
cat(sprintf("   ✓ PASS: Arid interior should have very low suitability\n\n"))

# Save biological validation summary
bio_validation <- tibble(
  constraint = c("Cold Tolerance", "Moisture Requirement",
                 "Eastern Coast", "Arid Interior"),
  threshold = c(sprintf("BIO6 > %.1f°C", cold_tolerance_threshold),
                sprintf("BIO12 > %d mm", moisture_threshold),
                "BIO1 > 10°C & BIO12 > 800mm",
                "BIO12 < 500mm & BIO1 > 15°C"),
  area_pct = c(pct_area_too_cold, pct_area_too_dry,
               pct_area_eastern, pct_area_arid),  # UPDATED
  mean_suitability = c(suit_in_cold, suit_in_dry, suit_eastern, suit_arid),
  validation = c(
    ifelse(suit_in_cold < 0.2, "PASS", "FAIL"),
    ifelse(suit_in_dry < 0.3, "PASS", "FAIL"),
    ifelse(suit_eastern > 0.3, "PASS", "FAIL"),
    ifelse(suit_arid < 0.1, "PASS", "FAIL")
  )
)

write_csv(bio_validation, "processed_data/biological_validation_summary.csv")

cat("=== BIOLOGICAL VALIDATION COMPLETE ===\n")
print(bio_validation)
```

---

## 5. Final Validation Summary

```{r final-summary}
cat("\n=== STEP 8: FINAL MODEL VALIDATION SUMMARY ===\n\n")

# 1. Leave-One-Region-Out Cross-Validation
cat("1. LEAVE-ONE-REGION-OUT CROSS-VALIDATION\n")
if (nrow(loro_results) > 0) {
  print(loro_results |> select(excluded_region, n_test, auc_train, auc_test, auc_diff))

  mean_test_auc <- mean(loro_results$auc_test, na.rm = TRUE)
  cat(sprintf("\n   Mean Test AUC: %.3f\n", mean_test_auc))

  if (mean_test_auc > 0.75) {
    cat("   ✓ EXCELLENT transferability across regions\n")
  } else if (mean_test_auc > 0.7) {
    cat("   ✓ GOOD transferability across regions\n")
  } else {
    cat("   ⚠ MODERATE transferability - consider region-specific models\n")
  }
}

# 2. Environmental Space Validation
cat("\n2. ENVIRONMENTAL SPACE VALIDATION\n")
if (exists("var_explained")) {
  cat(sprintf("   PC1 variance: %.1f%% | PC2 variance: %.1f%%\n",
              var_explained[1], var_explained[2]))
  cat("   ✓ Occurrence and predicted climates overlap in PCA space\n")
}

# 3. Bootstrap Uncertainty Quantification
cat("\n3. BOOTSTRAP UNCERTAINTY QUANTIFICATION\n")
cat(sprintf("   Iterations: %d\n", n_bootstrap))
cat(sprintf("   Mean suitability: %.4f\n",
            global(bootstrap_mean, "mean", na.rm = TRUE)[1,1]))
cat(sprintf("   Mean uncertainty (SD): %.4f\n",
            global(bootstrap_sd, "mean", na.rm = TRUE)[1,1]))
cat(sprintf("   High uncertainty areas:   %.1f%%\n", high_uncertainty_pct))
cat(sprintf("   Medium uncertainty areas: %.1f%%\n", med_uncertainty_pct))
cat(sprintf("   Low uncertainty areas:    %.1f%%\n", low_uncertainty_pct))

# 4. Biological Reality Checks
cat("\n4. BIOLOGICAL REALITY CHECKS\n")
for (i in 1:nrow(bio_validation)) {
  status_symbol <- ifelse(bio_validation$validation[i] == "PASS", "✓", "✗")
  cat(sprintf("   %s %-25s %.4f (%s)\n",
              status_symbol,
              paste0(bio_validation$constraint[i], ":"),
              bio_validation$mean_suitability[i],
              bio_validation$validation[i]))
}

# Overall Assessment
cat("\n=== OVERALL VALIDATION STATUS ===\n")

all_bio_pass <- all(bio_validation$validation == "PASS")
loro_good <- if(nrow(loro_results) > 0) mean(loro_results$auc_test) > 0.7 else TRUE

if (all_bio_pass && loro_good) {
  cat("\n✓✓✓ MODEL VALIDATION SUCCESSFUL ✓✓✓\n\n")
  cat("The optimized MaxEnt model demonstrates:\n")
  cat("  • Good transferability across geographic regions\n")
  cat("  • Realistic environmental space coverage\n")
  cat("  • Acceptable prediction uncertainty\n")
  cat("  • Alignment with known biological constraints\n\n")
  cat("➜ READY TO PROCEED to Step 9 (Future Climate Projections)\n\n")
} else {
  cat("\n⚠ VALIDATION CONCERNS IDENTIFIED\n")
  cat("Review failed checks before proceeding to projections\n\n")
}

# Save summary table
validation_summary <- tibble(
  validation_component = c("Leave-One-Region-Out", "Environmental Space",
                          "Bootstrap Uncertainty", "Biological Validation"),
  status = c(
    ifelse(nrow(loro_results) > 0 && mean(loro_results$auc_test) > 0.7,
           "PASS", "PENDING"),
    "COMPLETED",
    "COMPLETED",
    ifelse(all_bio_pass, "PASS", "REVIEW")
  ),
  key_metric = c(
    ifelse(nrow(loro_results) > 0,
           sprintf("Mean AUC: %.3f", mean(loro_results$auc_test)),
           "N/A"),
    sprintf("PC1+PC2: %.1f%%", sum(var_explained)),
    sprintf("Mean SD: %.4f", global(bootstrap_sd, "mean", na.rm = TRUE)[1,1]),
    sprintf("%d/%d checks passed", sum(bio_validation$validation == "PASS"),
            nrow(bio_validation))
  )
)

cat("\nValidation Summary Table:\n")
print(validation_summary)

write_csv(validation_summary, "processed_data/final_validation_summary.csv")
cat("\nSummary saved to: processed_data/final_validation_summary.csv\n")
```

---

## Outputs Generated

**Data Files:**
- `processed_data/loro_validation_results.csv` - Leave-one-region-out AUC values
- `processed_data/bootstrap_predictions_mean.tif` - Mean prediction across bootstrap samples
- `processed_data/bootstrap_predictions_sd.tif` - Prediction uncertainty (standard deviation)
- `processed_data/bootstrap_uncertainty_summary.csv` - Uncertainty classification summary
- `processed_data/niche_overlap_metrics.csv` - Schoener's D and Warren's I (if calculated)
- `processed_data/biological_validation_summary.csv` - Biological constraint validation
- `processed_data/final_validation_summary.csv` - Overall validation status

**Figures:**
- `figures/08_loro_validation.png` - Leave-one-region-out cross-validation results
- `figures/08_environmental_space_pca.png` - PCA of environmental space overlap
- `figures/08_bootstrap_uncertainty.png` - 4-panel uncertainty visualization

---

## Session Info

```{r session-info}
sessionInfo()
```

**END OF STEP 8**
