---
title: "Step 9: Future Climate Data Download"
author: "Species Distribution Modeling Pipeline"
date: "`r Sys.Date()`"
format:
  gfm:
    toc: true
    toc-depth: 3
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

## Overview

This script downloads future climate projection data from WorldClim 2.1 for projecting species distributions under climate change scenarios. We will download:

**Scenarios**: 12 total combinations
- **3 GCMs** (Global Climate Models):
  - MIROC6 (Japan)
  - MPI-ESM1-2-HR (Germany)
  - UKESM1-0-LL (UK)
- **2 SSPs** (Shared Socioeconomic Pathways):
  - SSP2-4.5 (Middle-of-the-road emissions)
  - SSP5-8.5 (High emissions/worst case)
- **2 Time Periods**:
  - 2041-2060 (Mid-century)
  - 2081-2100 (End-century)

**Resolution**: 2.5 minutes (matching current climate data)

**Expected Runtime**: 30-60 minutes (depending on download speeds)

**Data Source**: WorldClim 2.1 CMIP6 projections via geodata package

---

## Setup

```{r setup}
#| message: false
#| warning: false

# Set options
terra::terraOptions(memfrac = 0.8, tempdir = "temp")

# Core packages
library(terra)
library(sf)
library(tidyverse)
library(geodata)  # For downloading WorldClim future data

# Create directories
#dir.create("processed_data/future_climate", showWarnings = FALSE, recursive = TRUE)
#dir.create("temp", showWarnings = FALSE)


processed_data_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/processed_data/"

cat("✓ Setup complete\n")
```

---

## Configuration

Define the scenarios to download:

```{r config}
# GCMs to download (3 models for ensemble)
gcms <- c("MIROC6", "MPI-ESM1-2-HR", "UKESM1-0-LL")

# SSPs (emission scenarios)
ssps <- c("245", "585")  # SSP2-4.5 and SSP5-8.5

# Time periods
periods <- c("2041-2060", "2081-2100")

# Resolution (must match current climate: 2.5 minutes)
resolution <- 2.5

# Variables to match current model
selected_vars <- c("bio1", "bio6", "bio5", "bio12", "bio15", "bio3")

cat("=== Download Configuration ===\n")
cat(sprintf("GCMs: %s\n", paste(gcms, collapse = ", ")))
cat(sprintf("SSPs: SSP%s\n", paste(ssps, collapse = ", SSP")))
cat(sprintf("Periods: %s\n", paste(periods, collapse = ", ")))
cat(sprintf("Resolution: %.1f arcmin\n", resolution))
cat(sprintf("Variables: %s\n", paste(selected_vars, collapse = ", ")))
cat(sprintf("\nTotal scenarios: %d GCMs × %d SSPs × %d periods = %d\n",
            length(gcms), length(ssps), length(periods),
            length(gcms) * length(ssps) * length(periods)))
```

---

## Load Current Climate Reference

Load current climate extent to ensure future data matches:

```{r load-current}
# Load current climate data for reference
bioclim_current <- rast(paste0(processed_data_dir, "bioclim_selected.rds"))

# Get extent for cropping
current_extent <- ext(bioclim_current)

# Load Australia/NZ boundary for cropping
aus_nz_boundary <- st_read(paste0(processed_data_dir, "aus_nz_boundary.gpkg"), quiet = TRUE)

cat("✓ Current climate reference loaded\n")
cat(sprintf("  Extent: xmin=%.2f, xmax=%.2f, ymin=%.2f, ymax=%.2f\n",
            current_extent[1], current_extent[2],
            current_extent[3], current_extent[4]))
cat(sprintf("  Resolution: %.4f° (%.2f arcmin)\n",
            res(bioclim_current)[1],
            res(bioclim_current)[1] * 60))
cat(sprintf("  Variables: %s\n", paste(names(bioclim_current), collapse=", ")))
```

---

## Download Future Climate Data

Download all 12 scenario combinations:

```{r download-future}
# Only needs to be run once - comment out if already downloaded

# cat("\n=== Downloading Future Climate Data ===\n\n")

# # Initialize progress tracking
# download_log <- tibble(
#   gcm = character(),
#   ssp = character(),
#   period = character(),
#   status = character(),
#   file_path = character(),
#   n_layers = numeric(),
#   download_time_sec = numeric()
# )

# # Total iterations
# total_scenarios <- length(gcms) * length(ssps) * length(periods)
# current_scenario <- 0

# # Download loop
# for (gcm in gcms) {
#   for (ssp in ssps) {
#     for (period in periods) {
#       current_scenario <- current_scenario + 1

#       cat(sprintf("\n[%d/%d] Downloading: %s, SSP%s, %s\n",
#                   current_scenario, total_scenarios, gcm, ssp, period))

#       # Define file path
#       file_name <- sprintf("future_bioclim_%s_ssp%s_%s.tif",
#                           tolower(gsub("-", "_", gcm)),
#                           ssp,
#                           gsub("-", "_", period))
#       file_path <- file.path("processed_data/future_climate", file_name)

#       # Check if already downloaded
#       if (file.exists(file_path)) {
#         cat("  ✓ Already downloaded (skipping)\n")

#         # Load to check layers
#         existing_rast <- rast(file_path)

#         download_log <- download_log %>%
#           add_row(
#             gcm = gcm,
#             ssp = paste0("SSP", ssp),
#             period = period,
#             status = "cached",
#             file_path = file_path,
#             n_layers = nlyr(existing_rast),
#             download_time_sec = 0
#           )

#         next
#       }

#       # Download with timing
#       start_time <- Sys.time()

#       tryCatch({
#         # Download using geodata::cmip6_world
#         # Note: geodata expects ssp format as "245" not "ssp245"
#         future_data <- cmip6_world(
#           model = gcm,
#           ssp = ssp,
#           time = period,
#           var = "bioc",  # Bioclimatic variables
#           res = resolution,
#           path = "temp"
#         )

#         # Crop to Australia/NZ extent
#         future_data_cropped <- crop(future_data, vect(aus_nz_boundary))

#         # Select only the variables used in the model
#         # WorldClim naming: wc2.1_2.5m_bioc_GCMNAME_sspSSP_PERIOD_bio1
#         # Need to match bio1, bio3, bio5, bio6, bio12, bio15

#         # Get layer names and inspect
#         layer_names <- names(future_data_cropped)
#         cat(sprintf("  → Downloaded layers (%d total): %s ... %s\n",
#                     nlyr(future_data_cropped),
#                     paste(head(layer_names, 3), collapse=" "),
#                     tail(layer_names, 1)))

#         # Standardize layer names to simple bio1, bio2, etc.
#         # Handle multiple possible naming patterns:
#         if (grepl("wc2.1.*_\\d+$", layer_names[1])) {
#           # Pattern: wc2.1_2.5m_bioc_MIROC6_ssp585_2041-2060_1
#           # Extract the final number which is the bio variable number
#           bio_numbers <- as.numeric(gsub(".*_(\\d+)$", "\\1", layer_names))
#           names(future_data_cropped) <- paste0("bio", bio_numbers)
#         } else if (grepl("bio\\d+$", layer_names[1])) {
#           # Pattern: already ends with bio1, bio2, etc.
#           # Just extract the numbers
#           bio_numbers <- gsub(".*bio(\\d+)$", "\\1", layer_names)
#           names(future_data_cropped) <- paste0("bio", bio_numbers)
#         } else {
#           # Unknown pattern - just use sequential numbering
#           warning(sprintf("Unknown layer name pattern: %s", layer_names[1]))
#           names(future_data_cropped) <- paste0("bio", 1:nlyr(future_data_cropped))
#         }

#         cat(sprintf("  → Renamed to: %s\n",
#                     paste(names(future_data_cropped), collapse=", ")))

#         # Select only the variables we need
#         # Use tryCatch to handle subsetting errors
#         future_selected <- tryCatch({
#           future_data_cropped[[selected_vars]]
#         }, error = function(e) {
#           cat(sprintf("  ⚠ Error selecting variables: %s\n", e$message))
#           cat(sprintf("  Available: %s\n", paste(names(future_data_cropped), collapse=", ")))
#           cat(sprintf("  Requested: %s\n", paste(selected_vars, collapse=", ")))
#           stop("Layer selection failed - check variable names")
#         })

#         cat(sprintf("  → Selected %d variables: %s\n",
#                     nlyr(future_selected),
#                     paste(names(future_selected), collapse=", ")))

#         # Resample to match current climate resolution exactly
#         future_resampled <- resample(future_selected, bioclim_current,
#                                     method = "bilinear")

#         # Save
#         writeRaster(future_resampled, file_path, overwrite = TRUE)

#         end_time <- Sys.time()
#         elapsed_sec <- as.numeric(difftime(end_time, start_time, units = "secs"))

#         cat(sprintf("  ✓ Downloaded and processed (%.1f seconds)\n", elapsed_sec))
#         cat(sprintf("  ✓ Saved: %s\n", basename(file_path)))
#         cat(sprintf("  ✓ Layers: %d variables\n", nlyr(future_resampled)))

#         # Log success
#         download_log <- download_log %>%
#           add_row(
#             gcm = gcm,
#             ssp = paste0("SSP", ssp),
#             period = period,
#             status = "success",
#             file_path = file_path,
#             n_layers = nlyr(future_resampled),
#             download_time_sec = elapsed_sec
#           )

#       }, error = function(e) {
#         cat(sprintf("  ✗ ERROR: %s\n", e$message))

#         # Log failure
#         download_log <- download_log %>%
#           add_row(
#             gcm = gcm,
#             ssp = paste0("SSP", ssp),
#             period = period,
#             status = "failed",
#             file_path = NA_character_,
#             n_layers = 0,
#             download_time_sec = 0
#           )
#       })

#       # Small delay between downloads to be polite to servers
#       Sys.sleep(2)
#     }
#   }
# }

# # Save download log
# write_csv(download_log, paste0(processed_data_dir, "future_climate_download_log.csv"))

# cat("\n=== Download Summary ===\n")
# print(download_log)
```


```{r}
# Only needs to be run once - comment out if already downloaded

# # Retry the failed download
# gcm <- "UKESM1-0-LL"
# ssp <- "245"
# period <- "2081-2100"

# cat(sprintf("Retrying: %s, SSP%s, %s\n", gcm, ssp, period))

# file_name <- sprintf("future_bioclim_%s_ssp%s_%s.tif",
#                     tolower(gsub("-", "_", gcm)),
#                     ssp,
#                     gsub("-", "_", period))
# file_path <- file.path(paste0(processed_data_dir, "future_climate"), file_name)

# # Delete incomplete file if it exists
# if (file.exists(file_path)) {
#   file.remove(file_path)
#   cat("Removed incomplete file\n")
# }

# # Download with error handling
# start_time <- Sys.time()

# tryCatch({
#   future_data <- cmip6_world(
#     model = gcm,
#     ssp = ssp,
#     time = period,
#     var = "bioc",
#     res = resolution,
#     path = "temp"
#   )

#   # Crop to Australia/NZ extent
#   future_data_cropped <- crop(future_data, vect(aus_nz_boundary))

#   # Standardize layer names
#   layer_names <- names(future_data_cropped)
#   bio_numbers <- as.numeric(gsub(".*_(\\d+)$", "\\1", layer_names))
#   names(future_data_cropped) <- paste0("bio", bio_numbers)

#   cat(sprintf("  → Renamed to: %s\n", paste(names(future_data_cropped), collapse=", ")))

#   # Select required variables
#   future_selected <- future_data_cropped[[selected_vars]]

#   # Resample to match current climate
#   future_resampled <- resample(future_selected, bioclim_current, method = "bilinear")

#   # Save
#   writeRaster(future_resampled, file_path, overwrite = TRUE)

#   end_time <- Sys.time()
#   elapsed_sec <- as.numeric(difftime(end_time, start_time, units = "secs"))

#   cat(sprintf("  ✓ Downloaded and processed (%.1f seconds)\n", elapsed_sec))
#   cat(sprintf("  ✓ Saved: %s\n", basename(file_path)))

#   # Update download log
#   download_log <- download_log |>
#     filter(!(gcm == "UKESM1-0-LL" & ssp == "SSP245" & period == "2081-2100")) |>
#     add_row(
#       gcm = gcm,
#       ssp = paste0("SSP", ssp),
#       period = period,
#       status = "success",
#       file_path = file_path,
#       n_layers = nlyr(future_resampled),
#       download_time_sec = elapsed_sec
#     )

#   # Save updated log
#   write_csv(download_log, paste0(processed_data_dir, "future_climate_download_log.csv"))

# "future_climate_download_log.csv")

#   cat("\n✓ Download log updated\n")

# }, error = function(e) {
#   cat(sprintf("  ✗ ERROR: %s\n", e$message))
#   cat("  Check internet connection and try again\n")
# })
```

```{r}
# This has been edited to contain FULL PATHS - needed for rendering
download_log <- read.csv(paste0(processed_data_dir, "future_climate_download_log.csv"))
```

## Validate Downloaded Data

Check that all downloads succeeded and data integrity:

```{r validate-downloads}
# Not needed once data is downloaded

# cat("\n=== Validating Downloaded Data ===\n\n")

# # Count successes
# n_success <- sum(download_log$status %in% c("success", "cached"))
# n_failed <- sum(download_log$status == "failed")

# cat(sprintf("Successful downloads: %d/%d\n", n_success, total_scenarios))
# cat(sprintf("Failed downloads: %d/%d\n", n_failed, total_scenarios))

# if (n_failed > 0) {
#   cat("\n⚠️ WARNING: Some downloads failed!\n")
#   cat("Failed scenarios:\n")
#   failed <- download_log %>% filter(status == "failed")
#   print(failed %>% select(gcm, ssp, period))
# } else {
#   cat("\n✓ All downloads successful!\n")
# }

# # Validate data integrity
# cat("\n--- Data Integrity Checks ---\n")

# validation_results <- tibble(
#   scenario = character(),
#   extent_match = logical(),
#   resolution_match = logical(),
#   n_layers = numeric(),
#   layer_names = character()
# )

# for (i in 1:nrow(download_log)) {
#   if (download_log$status[i] %in% c("success", "cached")) {
#     file_path <- download_log$file_path[i]

#     # Load raster
#     future_rast <- rast(file_path)

#     # Check extent (within 0.01 degree tolerance)
#     # Compare each element of the extent separately
#     ext_future <- as.vector(ext(future_rast))
#     ext_current <- as.vector(ext(bioclim_current))
#     extent_match <- all(abs(ext_future - ext_current) < 0.01)

#     # Check resolution (within 0.0001 degree tolerance)
#     res_match <- all(abs(res(future_rast) - res(bioclim_current)) < 0.0001)

#     # Get layer info
#     n_layers <- nlyr(future_rast)
#     layer_names <- paste(names(future_rast), collapse = ", ")

#     # Store results
#     validation_results <- validation_results %>%
#       add_row(
#         scenario = sprintf("%s_%s_%s",
#                           download_log$gcm[i],
#                           download_log$ssp[i],
#                           download_log$period[i]),
#         extent_match = extent_match,
#         resolution_match = res_match,
#         n_layers = n_layers,
#         layer_names = layer_names
#       )
#   }
# }

# # Check if all validations passed
# all_valid <- all(validation_results$extent_match) &&
#              all(validation_results$resolution_match) &&
#              all(validation_results$n_layers == length(selected_vars))

# if (all_valid) {
#   cat("\n✓✓✓ ALL VALIDATION CHECKS PASSED ✓✓✓\n")
#   cat("  • All extents match current climate\n")
#   cat("  • All resolutions match current climate\n")
#   cat(sprintf("  • All scenarios have %d variables\n", length(selected_vars)))
# } else {
#   cat("\n⚠️ VALIDATION ISSUES DETECTED:\n")
#   if (!all(validation_results$extent_match)) {
#     cat("  ✗ Some extents do not match\n")
#   }
#   if (!all(validation_results$resolution_match)) {
#     cat("  ✗ Some resolutions do not match\n")
#   }
#   if (!all(validation_results$n_layers == length(selected_vars))) {
#     cat(sprintf("  ✗ Some scenarios do not have %d variables\n",
#                 length(selected_vars)))
#   }
# }

# # Save validation results
# write_csv(validation_results, paste0(processed_data_dir, "future_climate_validation.csv"))

# cat("\n--- Validation Summary ---\n")
# print(validation_results)
```

---

## Calculate Climate Change Deltas

Calculate the change (delta) between current and future climate for each scenario:

```{r climate-deltas}
cat("\n=== Calculating Climate Change Deltas ===\n\n")

# Count successes
n_success <- sum(download_log$status %in% c("success", "cached"))
n_failed <- sum(download_log$status == "failed")


# Initialize storage
delta_summary <- tibble(
  scenario = character(),
  gcm = character(),
  ssp = character(),
  period = character(),
  variable = character(),
  mean_current = numeric(),
  mean_future = numeric(),
  mean_delta = numeric(),
  mean_percent_change = numeric()
)

# Loop through successful downloads
for (i in 1:nrow(download_log)) {
  if (download_log$status[i] %in% c("success", "cached")) {
    scenario_name <- sprintf("%s_%s_%s",
                            download_log$gcm[i],
                            download_log$ssp[i],
                            download_log$period[i])

    cat(sprintf("Processing: %s\n", scenario_name))

    # Load future climate
    future_rast <- rast(download_log$file_path[i])

    # Calculate deltas for each variable
    for (var in selected_vars) {
      current_var <- bioclim_current[[var]]
      future_var <- future_rast[[var]]

      # Calculate means
      mean_current <- global(current_var, "mean", na.rm = TRUE)[1,1]
      mean_future <- global(future_var, "mean", na.rm = TRUE)[1,1]
      mean_delta <- mean_future - mean_current

      # Percent change (careful with temperature vs precipitation)
      # For temperature (bio1, bio5, bio6): use absolute change
      # For precipitation (bio12, bio15): use percent change
      if (var %in% c("bio12", "bio15")) {
        percent_change <- (mean_delta / mean_current) * 100
      } else {
        percent_change <- mean_delta  # Absolute change for temperature
      }

      # Store results
      delta_summary <- delta_summary %>%
        add_row(
          scenario = scenario_name,
          gcm = download_log$gcm[i],
          ssp = download_log$ssp[i],
          period = download_log$period[i],
          variable = var,
          mean_current = mean_current,
          mean_future = mean_future,
          mean_delta = mean_delta,
          mean_percent_change = percent_change
        )
    }
  }
}

# Save delta summary
write_csv(delta_summary, paste0(processed_data_dir, "climate_change_deltas.csv"))

cat("\n✓ Climate change deltas calculated\n")
cat(sprintf("  Total calculations: %d scenarios × %d variables = %d\n",
            n_success, length(selected_vars), nrow(delta_summary)))
```

---

## Visualize Climate Change Projections

Create summary plots of projected climate changes:

```{r visualize-deltas}
# Temperature changes (bio1, bio5, bio6)
p_temp <- delta_summary %>%
  filter(variable %in% c("bio1", "bio5", "bio6")) %>%
  mutate(
    variable = recode(
      variable,
      "bio1" = "Annual Mean Temp",
      "bio5" = "Max Temp Warmest Month",
      "bio6" = "Min Temp Coldest Month"
    ),
    period_label = recode(
      period,
      "2041-2060" = "Mid-century\n(2041-2060)",
      "2081-2100" = "End-century\n(2081-2100)"
    )
  ) %>%
  ggplot(aes(x = gcm, y = mean_delta, fill = ssp)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  facet_grid(variable ~ period_label, scales = "free_y") +
  scale_fill_manual(
    values = c("SSP245" = "#4575b4", "SSP585" = "#d73027"),
    labels = c("SSP245" = "SSP2-4.5 (Moderate)", "SSP585" = "SSP5-8.5 (High)")
  ) +
  labs(
    title = "Projected Temperature Changes",
    subtitle = "Change from current climate (°C)",
    x = "Global Climate Model",
    y = "Temperature Change (°C)",
    fill = "Emission Scenario"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

ggsave(
  "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/figures/09_temperature_projections.png",
  p_temp,
  width = 12,
  height = 10,
  dpi = 300
)

print(p_temp)

# Precipitation changes (bio12, bio15)
p_precip <- delta_summary %>%
  filter(variable %in% c("bio12", "bio15")) %>%
  mutate(
    variable = recode(
      variable,
      "bio12" = "Annual Precipitation",
      "bio15" = "Precipitation Seasonality"
    ),
    period_label = recode(
      period,
      "2041-2060" = "Mid-century\n(2041-2060)",
      "2081-2100" = "End-century\n(2081-2100)"
    )
  ) %>%
  ggplot(aes(x = gcm, y = mean_percent_change, fill = ssp)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  facet_grid(variable ~ period_label, scales = "free_y") +
  scale_fill_manual(
    values = c("SSP245" = "#4575b4", "SSP585" = "#d73027"),
    labels = c("SSP245" = "SSP2-4.5 (Moderate)", "SSP585" = "SSP5-8.5 (High)")
  ) +
  labs(
    title = "Projected Precipitation Changes",
    subtitle = "Percent change from current climate",
    x = "Global Climate Model",
    y = "Change (%)",
    fill = "Emission Scenario"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  )

ggsave(
  "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/figures/09_precipitation_projections.png",
  p_precip,
  width = 12,
  height = 8,
  dpi = 300
)

print(p_precip)

# Summary table of mean changes by SSP and period
summary_by_ssp <- delta_summary %>%
  group_by(ssp, period, variable) %>%
  summarise(
    mean_delta = mean(mean_delta),
    sd_delta = sd(mean_delta),
    .groups = "drop"
  ) %>%
  arrange(variable, ssp, period)

cat("\n=== Mean Climate Changes by SSP and Period ===\n")
print(summary_by_ssp, n = Inf)

write_csv(
  summary_by_ssp,
  paste0(processed_data_dir, "climate_change_summary_by_ssp.csv")
)
```

---

## Final Summary

```{r final-summary}
cat("\n=== STEP 9: FUTURE CLIMATE DATA DOWNLOAD - COMPLETE ===\n\n")

cat("CONFIGURATION\n")
cat(sprintf("  GCMs: %d models\n", length(gcms)))
cat("    • MIROC6 (Japan)\n")
cat("    • MPI-ESM1-2-HR (Germany)\n")
cat("    • UKESM1-0-LL (UK)\n\n")

cat("  Emission Scenarios: 2\n")
cat("    • SSP2-4.5 (Moderate emissions)\n")
cat("    • SSP5-8.5 (High emissions)\n\n")

cat("  Time Periods: 2\n")
cat("    • 2041-2060 (Mid-century)\n")
cat("    • 2081-2100 (End-century)\n\n")

cat(sprintf(
  "  Variables: %d (matching current model)\n",
  length(selected_vars)
))
cat("    • bio1, bio3, bio5, bio6, bio12, bio15\n\n")

cat("CLIMATE CHANGE SUMMARY\n")

# Calculate summary statistics
temp_change_2050_ssp245 <- delta_summary %>%
  filter(variable == "bio1", ssp == "SSP245", period == "2041-2060") %>%
  summarise(mean = mean(mean_delta)) %>%
  pull(mean)

temp_change_2090_ssp585 <- delta_summary %>%
  filter(variable == "bio1", ssp == "SSP585", period == "2081-2100") %>%
  summarise(mean = mean(mean_delta)) %>%
  pull(mean)

precip_change_2050_ssp245 <- delta_summary %>%
  filter(variable == "bio12", ssp == "SSP245", period == "2041-2060") %>%
  summarise(mean = mean(mean_percent_change)) %>%
  pull(mean)

precip_change_2090_ssp585 <- delta_summary %>%
  filter(variable == "bio12", ssp == "SSP585", period == "2081-2100") %>%
  summarise(mean = mean(mean_percent_change)) %>%
  pull(mean)

cat("  Temperature (Annual Mean) Changes:\n")
cat(sprintf("    Mid-century SSP2-4.5:  +%.2f°C\n", temp_change_2050_ssp245))
cat(sprintf("    End-century SSP5-8.5:  +%.2f°C\n\n", temp_change_2090_ssp585))

cat("  Precipitation (Annual) Changes:\n")
cat(sprintf("    Mid-century SSP2-4.5:  %+.1f%%\n", precip_change_2050_ssp245))
cat(sprintf(
  "    End-century SSP5-8.5:  %+.1f%%\n\n",
  precip_change_2090_ssp585
))

cat("OUTPUTS GENERATED\n\n")
cat("  Data Files:\n")
cat(sprintf("    • %d future climate rasters (.tif)\n", n_success))
cat("    • future_climate_download_log.csv\n")
cat("    • future_climate_validation.csv\n")
cat("    • climate_change_deltas.csv\n")
cat("    • climate_change_summary_by_ssp.csv\n\n")

cat("  Figures:\n")
cat("    • 09_temperature_projections.png\n")
cat("    • 09_precipitation_projections.png\n\n")
```

---

## Outputs Generated

**Data Files** (in `processed_data/future_climate/`):
- 12 future climate rasters: `future_bioclim_[gcm]_ssp[###]_[period].tif`
- `future_climate_download_log.csv` - Download tracking and status
- `future_climate_validation.csv` - Data integrity validation results
- `climate_change_deltas.csv` - Detailed change metrics for all scenarios
- `climate_change_summary_by_ssp.csv` - Aggregated changes by SSP and period

**Figures**:
- `figures/09_temperature_projections.png` - Temperature changes by GCM, SSP, period
- `figures/09_precipitation_projections.png` - Precipitation changes by GCM, SSP, period

---

## Session Info

```{r session-info}
sessionInfo()
```

**END OF STEP 9**
