---
title: "Step 10: Future Climate Projections"
author: "Species Distribution Modeling Pipeline"
date: "`r Sys.Date()`"
format:
  gfm:
    toc: true
    toc-depth: 3
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
---

## Overview

This script projects the optimized MaxEnt model to future climate scenarios and analyzes habitat suitability changes.

**What This Script Does**:
1. Loads the optimized MaxEnt model (from Step 7)
2. Projects to all 12 future climate scenarios (from Step 9)
3. Creates ensemble predictions (mean and SD across GCMs)
4. Calculates habitat suitability change (future - current)
5. Identifies areas of gain, loss, and stability
6. Quantifies suitable habitat area by scenario

**Scenarios**: 12 total combinations
- 3 GCMs: MIROC6, MPI-ESM1-2-HR, UKESM1-0-LL
- 2 SSPs: SSP2-4.5 (moderate), SSP5-8.5 (high emissions)
- 2 time periods: 2041-2060 (mid-century), 2081-2100 (end-century)

**Expected Runtime**: 15-30 minutes

---

## Setup

```{r setup}
#| message: false
#| warning: false

# Set options
options(java.parameters = "-Xmx12g")
terra::terraOptions(memfrac = 0.8, tempdir = "temp")

# Core packages
library(terra)
library(tidyverse)
library(predicts)
library(sf)

# Visualization
library(ggplot2)
library(patchwork)
library(viridis)
library(tidyterra)

processed_data_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/processed_data/"

# Create directories
#dir.create("outputs/future_projections", showWarnings = FALSE, recursive = TRUE)
#dir.create("figures", showWarnings = FALSE)
```

---

## Load Current Climate and Model

```{r load-model}
# Loading Model and Current Climate

# Load optimized MaxEnt model
maxent_model <- readRDS(paste0(processed_data_dir, "maxent_optimized_model.rds"))

# Load current climate prediction for comparison
current_prediction <- rast(paste0(processed_data_dir, "maxent_optimized_prediction_aus.tif"))

cat("✓ Optimized MaxEnt model loaded\n")
cat("✓ Current prediction loaded\n")
cat(sprintf("  Extent: xmin=%.2f, xmax=%.2f, ymin=%.2f, ymax=%.2f\n",
            ext(current_prediction)[1], ext(current_prediction)[2],
            ext(current_prediction)[3], ext(current_prediction)[4]))
cat(sprintf("  Resolution: %.4f°\n", res(current_prediction)[1]))
```

---

## Project to Future Scenarios

Loop through all 12 future climate scenarios and generate predictions:

```{r project-future}
# Projecting to Future Scenarios

# Configuration
gcms <- c("MIROC6", "MPI-ESM1-2-HR", "UKESM1-0-LL")
ssps <- c("245", "585")
periods <- c("2041-2060", "2081-2100")

# Create output directory
#dir.create("processed_data/future_projections", showWarnings = FALSE, recursive = TRUE)

# Initialize tracking
projection_log <- tibble(
  gcm = character(),
  ssp = character(),
  period = character(),
  status = character(),
  mean_suitability = numeric(),
  suitable_area_km2 = numeric(),
  processing_time_sec = numeric()
)

# Total scenarios
total_scenarios <- length(gcms) * length(ssps) * length(periods)
current_scenario <- 0

# Storage for all predictions
all_predictions <- list()

# Projection loop
for (gcm in gcms) {
  for (ssp in ssps) {
    for (period in periods) {
      current_scenario <- current_scenario + 1

      scenario_name <- sprintf("%s_ssp%s_%s",
                               tolower(gsub("-", "_", gcm)),
                               ssp,
                               gsub("-", "_", period))

cat(sprintf("[%d/%d] Projecting: %s, SSP%s, %s\n",
                  current_scenario, total_scenarios, gcm, ssp, period))

      start_time <- Sys.time()

      tryCatch({
        # Load future climate data
        future_climate_file <- sprintf(
          paste0(
            processed_data_dir,
            "future_climate/future_bioclim_%s.tif"),
          scenario_name
        )

        if (!file.exists(future_climate_file)) {
          stop(sprintf("Future climate file not found: %s", future_climate_file))
        }

        future_climate <- rast(future_climate_file)

        # Predict with MaxEnt model
        future_prediction <- predict(
          maxent_model,
          future_climate,
          args = c('outputformat=logistic')
        )

        # Calculate statistics
        mean_suit <- global(future_prediction, "mean", na.rm = TRUE)[1,1]

        # Calculate suitable area (threshold = 0.5)
        suitable_cells <- sum(values(future_prediction > 0.5), na.rm = TRUE)
        cell_area_km2 <- prod(res(future_prediction)) * 111 * 111  # Approx km²
        suitable_area <- suitable_cells * cell_area_km2

        # Save individual prediction
        output_file <- sprintf(
          paste0(processed_data_dir, "future_projections/%s_prediction.tif"),
          scenario_name
        )
        writeRaster(future_prediction, output_file, overwrite = TRUE)

        # Store in list for ensemble calculation
        all_predictions[[scenario_name]] <- future_prediction

        end_time <- Sys.time()
        elapsed_sec <- as.numeric(difftime(end_time, start_time, units = "secs"))

        cat(sprintf("  ✓ Mean suitability: %.4f\n", mean_suit))
        cat(sprintf("  ✓ Suitable area: %s km²\n", scales::comma(suitable_area, accuracy = 1)))
        cat(sprintf("  ✓ Saved: %s (%.1f sec)\n", basename(output_file), elapsed_sec))

        # Log success
        projection_log <- projection_log %>%
          add_row(
            gcm = gcm,
            ssp = paste0("SSP", ssp),
            period = period,
            status = "success",
            mean_suitability = mean_suit,
            suitable_area_km2 = suitable_area,
            processing_time_sec = elapsed_sec
          )

      }, error = function(e) {
        cat(sprintf("  ✗ ERROR: %s\n", e$message))

        # Log failure
        projection_log <- projection_log %>%
          add_row(
            gcm = gcm,
            ssp = paste0("SSP", ssp),
            period = period,
            status = "failed",
            mean_suitability = NA_real_,
            suitable_area_km2 = NA_real_,
            processing_time_sec = 0
          )
      })
    }
  }
}

# Save projection log
write_csv(projection_log, paste0(processed_data_dir, "future_projections/projection_log.csv"))

cat("\n=== Projection Summary ===\n")
cat(sprintf("Total scenarios: %d\n", total_scenarios))
cat(sprintf("Successful: %d\n", sum(projection_log$status == "success")))
cat(sprintf("Failed: %d\n", sum(projection_log$status == "failed")))

if (sum(projection_log$status == "failed") > 0) {
  cat("\n⚠ Failed scenarios:\n")
  print(projection_log %>% filter(status == "failed") %>%
          select(gcm, ssp, period))
}
```

---

## Create Ensemble Predictions

Calculate ensemble means and standard deviations across GCMs for each SSP × period combination:

```{r ensemble}
# Creating Ensemble Predictions

# Get unique SSP × period combinations
ssp_period_combos <- expand.grid(
  ssp = ssps,
  period = periods,
  stringsAsFactors = FALSE
)

ensemble_stats <- tibble(
  ssp = character(),
  period = character(),
  mean_suitability = numeric(),
  sd_suitability = numeric(),
  suitable_area_km2 = numeric()
)

for (i in 1:nrow(ssp_period_combos)) {
  ssp <- ssp_period_combos$ssp[i]
  period <- ssp_period_combos$period[i]

  cat(sprintf("Creating ensemble for SSP%s, %s\n", ssp, period))

  # Get all predictions for this SSP × period
  scenario_pattern <- sprintf("ssp%s_%s", ssp, gsub("-", "_", period))
  matching_scenarios <- names(all_predictions)[grepl(scenario_pattern, names(all_predictions))]

  if (length(matching_scenarios) == 0) {
    cat("  ⚠ No predictions found for this combination\n")
    next
  }

  # Stack predictions for this SSP × period
  scenario_stack <- rast(lapply(matching_scenarios, function(s) all_predictions[[s]]))

  # Calculate ensemble mean and SD
  ensemble_mean <- mean(scenario_stack, na.rm = TRUE)
  ensemble_sd <- stdev(scenario_stack, na.rm = TRUE)

  # Save ensemble predictions
  ensemble_name <- sprintf("ensemble_ssp%s_%s", ssp, gsub("-", "_", period))

  writeRaster(
    ensemble_mean,
    sprintf(paste0(processed_data_dir, "future_projections/%s_mean.tif"), ensemble_name),
    overwrite = TRUE
  )

  writeRaster(
    ensemble_sd,
    sprintf(paste0(processed_data_dir, "future_projections/%s_sd.tif"), ensemble_name),
    overwrite = TRUE
  )

  # Calculate statistics
  mean_suit <- global(ensemble_mean, "mean", na.rm = TRUE)[1,1]
  sd_suit <- global(ensemble_sd, "mean", na.rm = TRUE)[1,1]
  suitable_cells <- sum(values(ensemble_mean > 0.5), na.rm = TRUE)
  cell_area_km2 <- prod(res(ensemble_mean)) * 111 * 111
  suitable_area <- suitable_cells * cell_area_km2

  cat(sprintf("  ✓ Mean suitability: %.4f (SD: %.4f)\n", mean_suit, sd_suit))
  cat(sprintf("  ✓ Suitable area: %s km²\n", scales::comma(suitable_area, accuracy = 1)))

  ensemble_stats <- ensemble_stats %>%
    add_row(
      ssp = paste0("SSP", ssp),
      period = period,
      mean_suitability = mean_suit,
      sd_suitability = sd_suit,
      suitable_area_km2 = suitable_area
    )
}

write_csv(ensemble_stats, paste0(processed_data_dir, "future_projections/ensemble_statistics.csv"))
```

---

## Calculate Change from Current

For each ensemble prediction, calculate change in habitat suitability:

```{r calculate-change}
# Calculating Habitat Suitability Change

change_stats <- tibble(
  ssp = character(),
  period = character(),
  mean_change = numeric(),
  area_gain_km2 = numeric(),
  area_loss_km2 = numeric(),
  area_stable_km2 = numeric()
)

# Cell area
cell_area_km2 <- prod(res(current_prediction)) * 111 * 111

for (i in 1:nrow(ssp_period_combos)) {
  ssp <- ssp_period_combos$ssp[i]
  period <- ssp_period_combos$period[i]

  ensemble_name <- sprintf("ensemble_ssp%s_%s", ssp, gsub("-", "_", period))
  ensemble_file <- sprintf(paste0(processed_data_dir, "future_projections/%s_mean.tif"), ensemble_name)

  if (!file.exists(ensemble_file)) {
    cat(sprintf("Skipping SSP%s %s (ensemble not found)\n", ssp, period))
    next
  }

  cat(sprintf("Calculating change for SSP%s, %s\n", ssp, period))

  # Load ensemble prediction
  future_ensemble <- rast(ensemble_file)

  # Ensure extents match by cropping and resampling to current_prediction
  future_ensemble <- crop(future_ensemble, current_prediction)
  future_ensemble <- resample(future_ensemble, current_prediction, method = "bilinear")

  # Calculate change (future - current)
  change_raster <- future_ensemble - current_prediction

  # Save change raster
  writeRaster(
    change_raster,
    sprintf(paste0(processed_data_dir, "future_projections/change_ssp%s_%s.tif"),
            ssp, gsub("-", "_", period)),
    overwrite = TRUE
  )

  # Calculate statistics
  mean_change <- global(change_raster, "mean", na.rm = TRUE)[1,1]

  # Classify change (using 0.5 threshold for suitable habitat)
  current_suitable <- current_prediction > 0.5
  future_suitable <- future_ensemble > 0.5

  # Gain: not suitable → suitable
  gain <- (!current_suitable) & future_suitable
  area_gain <- sum(values(gain), na.rm = TRUE) * cell_area_km2

  # Loss: suitable → not suitable
  loss <- current_suitable & (!future_suitable)
  area_loss <- sum(values(loss), na.rm = TRUE) * cell_area_km2

  # Stable: suitable → suitable
  stable <- current_suitable & future_suitable
  area_stable <- sum(values(stable), na.rm = TRUE) * cell_area_km2

  cat(sprintf("  Mean change: %+.4f\n", mean_change))
  cat(sprintf("  Area gain: %s km²\n", scales::comma(area_gain, accuracy = 1)))
  cat(sprintf("  Area loss: %s km²\n", scales::comma(area_loss, accuracy = 1)))
  cat(sprintf("  Area stable: %s km²\n", scales::comma(area_stable, accuracy = 1)))

  change_stats <- change_stats %>%
    add_row(
      ssp = paste0("SSP", ssp),
      period = period,
      mean_change = mean_change,
      area_gain_km2 = area_gain,
      area_loss_km2 = area_loss,
      area_stable_km2 = area_stable
    )
}

write_csv(change_stats, paste0(processed_data_dir, "future_projections/change_statistics.csv"))
```

---

## Summary Table: Suitable Habitat Area

```{r summary-table}
cat("\n=== Suitable Habitat Area by Scenario ===\n\n")

# Add current prediction to summary
current_suitable_cells <- sum(values(current_prediction > 0.5), na.rm = TRUE)
current_area <- current_suitable_cells * cell_area_km2

# Combine current and future
area_summary <- tibble(
  scenario = "Current (baseline)",
  ssp = "Current",
  period = "2000-2020",
  suitable_area_km2 = current_area,
  change_from_current_km2 = 0,
  change_from_current_pct = 0
) %>%
  bind_rows(
    ensemble_stats %>%
      mutate(
        scenario = paste(ssp, period),
        change_from_current_km2 = suitable_area_km2 - current_area,
        change_from_current_pct = (suitable_area_km2 - current_area) / current_area * 100
      ) %>%
      select(scenario, ssp, period, suitable_area_km2,
             change_from_current_km2, change_from_current_pct)
  )

# Print table
print(area_summary, n = 20)

# Save
write_csv(
  area_summary,
  paste0(
    processed_data_dir,
    "future_projections/suitable_area_summary.csv"
    )
  )
```

---

## Visualization: Ensemble Predictions

Create multi-panel figure showing ensemble predictions for all scenarios:

```{r viz-ensemble}
# Creating Ensemble Prediction Visualizations

# Define Australia/NZ extent for cropping
aus_nz_xlim <- c(110, 180)
aus_nz_ylim <- c(-48, -10)

# Load all ensemble predictions
ensemble_plots <- list()

for (i in 1:nrow(ssp_period_combos)) {
  ssp <- ssp_period_combos$ssp[i]
  period <- ssp_period_combos$period[i]

  ensemble_name <- sprintf("ensemble_ssp%s_%s", ssp, gsub("-", "_", period))
  ensemble_file <- sprintf(
    paste0(
      processed_data_dir,
      "future_projections/%s_mean.tif"
      ),
    ensemble_name
    )

  if (!file.exists(ensemble_file)) next

  ensemble_pred <- rast(ensemble_file)

  # Create plot with coord_sf limits
  plot_title <- sprintf("SSP%s, %s", ssp, period)

  p <- ggplot() +
    geom_spatraster(data = ensemble_pred) +
    scale_fill_gradientn(
      colors = c("lightgrey", "yellow", "orange", "red", "darkred"),
      na.value = "transparent",
      limits = c(0, 1),
      labels = scales::percent,
      name = "Habitat\nSuitability"
    ) +
    labs(title = plot_title) +
    coord_sf(xlim = aus_nz_xlim, ylim = aus_nz_ylim, expand = FALSE) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
      axis.title = element_blank(),
      panel.grid = element_line(color = "gray90", linewidth = 0.3)
    )

  ensemble_plots[[ensemble_name]] <- p
}

# Add current prediction for comparison
p_current <- ggplot() +
  geom_spatraster(data = current_prediction) +
  scale_fill_gradientn(
    colors = c("lightgrey", "yellow", "orange", "red", "darkred"),
    na.value = "transparent",
    limits = c(0, 1),
    labels = scales::percent,
    name = "Habitat\nSuitability"
  ) +
  labs(title = "Current (Baseline)") +
  coord_sf(xlim = aus_nz_xlim, ylim = aus_nz_ylim, expand = FALSE) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
    axis.title = element_blank(),
    panel.grid = element_line(color = "gray90", linewidth = 0.3)
  )

# Combine all plots
combined <- (p_current + ensemble_plots[[1]]) /
            (ensemble_plots[[2]] + ensemble_plots[[3]]) /
            (ensemble_plots[[4]] | plot_spacer())

print(combined)

ggsave(
  "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/figures/10_ensemble_predictions_all_scenarios.png",
       combined, width = 14, height = 12, dpi = 300)

```

---

## Visualization: Change Maps

Create change maps showing gain/loss/stable areas:

```{r viz-change}
cat("\n=== Creating Change Visualizations ===\n\n")

change_plots <- list()

for (i in 1:nrow(ssp_period_combos)) {
  ssp <- ssp_period_combos$ssp[i]
  period <- ssp_period_combos$period[i]

  change_file <- sprintf(
    paste0(processed_data_dir, "future_projections/change_ssp%s_%s.tif"),
    ssp,
    gsub("-", "_", period)
  )

  if (!file.exists(change_file)) {
    next
  }

  change_raster <- rast(change_file)

  # Create plot
  plot_title <- sprintf("Change: SSP%s, %s", ssp, period)

  p <- ggplot() +
    geom_spatraster(data = change_raster) +
    scale_fill_gradient2(
      low = "darkred",
      mid = "grey95",
      high = "darkblue",
      midpoint = 0,
      na.value = "transparent",
      limits = c(-1, 1),
      name = "Change in\nSuitability"
    ) +
    labs(title = plot_title) +
    coord_sf() +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", size = 11, hjust = 0.5),
      axis.title = element_blank(),
      panel.grid = element_line(color = "gray90", linewidth = 0.3)
    )

  change_plots[[i]] <- p
}

# Combine change plots
change_combined <- wrap_plots(change_plots, ncol = 2)

print(change_combined)

ggsave(
  "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/figures/10_change_maps_all_scenarios.png",
  change_combined,
  width = 12,
  height = 10,
  dpi = 300
)

cat("✓ Change maps saved\n")
```

---

## Final Summary

```{r final-summary}
cat("\n\n")
cat("╔════════════════════════════════════════════════════════════════╗\n")
cat("║         STEP 10: FUTURE PROJECTIONS - SUMMARY                ║\n")
cat("╠════════════════════════════════════════════════════════════════╣\n")

# Projection status
cat("║ 1. FUTURE PROJECTIONS                                         ║\n")
cat("╠════════════════════════════════════════════════════════════════╣\n")
cat(sprintf(
  "║   Scenarios projected: %d/%d                                    ║\n",
  sum(projection_log$status == "success"),
  total_scenarios
))
cat(sprintf(
  "║   Ensemble predictions: %d (by SSP × period)                  ║\n",
  nrow(ensemble_stats)
))

# Area changes
cat("╠════════════════════════════════════════════════════════════════╣\n")
cat("║ 2. SUITABLE HABITAT AREA                                      ║\n")
cat("╠════════════════════════════════════════════════════════════════╣\n")
cat(sprintf(
  "║   Current baseline: %s km²                          ║\n",
  format(current_area, big.mark = ",", scientific = FALSE)
))

for (i in 1:nrow(area_summary)) {
  if (area_summary$scenario[i] == "Current (baseline)") {
    next
  }

  cat(sprintf(
    "║   %-20s %s km² (%+.1f%%)              ║\n",
    paste0(area_summary$scenario[i], ":"),
    format(
      area_summary$suitable_area_km2[i],
      big.mark = ",",
      scientific = FALSE
    ),
    area_summary$change_from_current_pct[i]
  ))
}

# Change patterns
cat("╠════════════════════════════════════════════════════════════════╣\n")
cat("║ 3. HABITAT CHANGE PATTERNS                                    ║\n")
cat("╠════════════════════════════════════════════════════════════════╣\n")

for (i in 1:nrow(change_stats)) {
  cat(sprintf(
    "║   %s %s:                                       ║\n",
    change_stats$ssp[i],
    change_stats$period[i]
  ))
  cat(sprintf(
    "║     Gain: %s km² | Loss: %s km²               ║\n",
    format(change_stats$area_gain_km2[i], big.mark = ",", scientific = FALSE),
    format(change_stats$area_loss_km2[i], big.mark = ",", scientific = FALSE)
  ))
}

cat("╠════════════════════════════════════════════════════════════════╣\n")
cat("║ OUTPUTS GENERATED                                             ║\n")
cat("╠════════════════════════════════════════════════════════════════╣\n")
cat("║   • 12 individual scenario predictions                        ║\n")
cat("║   • 4 ensemble means (by SSP × period)                        ║\n")
cat("║   • 4 ensemble SDs (uncertainty)                              ║\n")
cat("║   • 4 change rasters (future - current)                       ║\n")
cat("║   • Summary tables (CSV)                                      ║\n")
cat("║   • Visualization figures (PNG)                               ║\n")
cat("╠════════════════════════════════════════════════════════════════╣\n")
cat("║                                                                ║\n")
cat("║   ✓✓✓ FUTURE PROJECTIONS COMPLETE ✓✓✓                        ║\n")
cat("║                                                                ║\n")
cat("║   ➜ READY TO PROCEED to Step 11 (Visualization & Results)     ║\n")
cat("║                                                                ║\n")
cat("╚════════════════════════════════════════════════════════════════╝\n\n")
```

---

## Outputs Generated

**Prediction Rasters**:
- `processed_data/future_projections/*_prediction.tif` (12 individual scenarios)
- `processed_data/future_projections/ensemble_*_mean.tif` (4 ensemble means)
- `processed_data/future_projections/ensemble_*_sd.tif` (4 ensemble SDs)
- `processed_data/future_projections/change_*.tif` (4 change rasters)

**Summary Tables**:
- `processed_data/future_projections/projection_log.csv`
- `processed_data/future_projections/ensemble_statistics.csv`
- `processed_data/future_projections/change_statistics.csv`
- `processed_data/future_projections/suitable_area_summary.csv`

**Figures**:
- `figures/10_ensemble_predictions_all_scenarios.png`
- `figures/10_change_maps_all_scenarios.png`

---

## Session Info

```{r session-info}
sessionInfo()
```

**END OF STEP 10**
