---
title: "Step 1: Occurrence Data — Revised"
author: "Alexander W. Gofton"
date: today
format: gfm
---

## Overview

This script revises the occurrence data from the original pipeline (v1) with two key changes:

1. **Remove records with missing coordinates** (1 record with NA lon/lat).
2. **Geographically stratified thinning** — cap each region at ~35% of total records to reduce the dominance of East Asian occurrences (previously 61% of all data).

We start from the already-thinned v1 dataset (614 records after 10 km spatial thinning) rather than re-running the full data assembly pipeline.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```

## 1. Load existing thinned occurrence data

The v1 pipeline produced a thinned dataset of 615 records (including 1 with NA coordinates) stored with all 19 WorldClim bioclimatic variables already extracted.

```{r}
#| label: load-data

# Paths — adjust base_dir to match your local setup
base_dir <- here::here()
processed_v1 <- file.path(base_dir, "processed_data")
processed_v2 <- file.path(base_dir, "processed_data_2")
figures_dir  <- file.path(base_dir, "figures_2")

# Load the v1 thinned occurrences (615 rows, 1 has NA coords)
occ_raw <- read_csv(
  file.path(processed_v1, "occurrences_thinned.csv"),
  show_col_types = FALSE
)

sprintf("Loaded %d records from v1 thinned dataset", nrow(occ_raw))
```

## 2. Remove records with missing coordinates

```{r}
#| label: remove-na

n_before <- nrow(occ_raw)

occ_clean <- occ_raw %>%
  filter(!is.na(lon), !is.na(lat))

sprintf(
  "Removed %d record(s) with missing coordinates: %d -> %d",
  n_before - nrow(occ_clean), n_before, nrow(occ_clean)
)
```

## 3. Assign geographic regions

We assign each record to one of three regions based on its coordinates:

- **East Asia**: China, Japan, Korea (N & S), Russia (Far East), Taiwan
- **Oceania**: Australia, New Zealand, Pacific Islands
- **North America**: USA

Records not falling into these categories are assigned to "Other". We use a spatial join with Natural Earth country boundaries.

```{r}
#| label: assign-regions

# Get world country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf") %>%
  select(name, iso_a3, continent, subregion, geometry)

# Convert occurrences to sf points
occ_sf <- st_as_sf(occ_clean, coords = c("lon", "lat"), crs = 4326)

# Spatial join to get country for each point
occ_with_country <- st_join(occ_sf, world, join = st_intersects)

# Define region mapping based on country/subregion
east_asia_countries <- c(
  "China", "Japan", "South Korea", "North Korea",
  "Taiwan", "Russia"    # Russian records in our data are Far East
)

oceania_countries <- c(
  "Australia", "New Zealand", "Fiji", "Tonga",
  "New Caledonia", "Vanuatu", "Samoa"
)

north_america_countries <- c("United States of America")

occ_with_region <- occ_with_country %>%
  mutate(
    region = case_when(
      name %in% east_asia_countries ~ "East_Asia",
      name %in% oceania_countries   ~ "Oceania",
      name %in% north_america_countries ~ "North_America",
      TRUE ~ "Other"
    )
  )

# Recover original lon/lat columns (lost during st_as_sf conversion)
coords <- st_coordinates(occ_with_region)
occ_with_region <- occ_with_region %>%
  st_drop_geometry() %>%
  mutate(lon = coords[, 1], lat = coords[, 2])

# Regional breakdown before stratified thinning
region_counts_before <- occ_with_region %>%
  count(region, name = "n_records") %>%
  mutate(pct = round(100 * n_records / sum(n_records), 1))

region_counts_before
```

## 4. Handle unassigned records

Some points may fall just offshore (small coordinate imprecisions) and miss the country polygons. Let's check and assign them using nearest-country logic.

```{r}
#| label: fix-unassigned

n_other <- sum(occ_with_region$region == "Other")

if (n_other > 0) {
  # For unassigned points, find the nearest country
  unassigned_idx <- which(occ_with_region$region == "Other")

  unassigned_sf <- st_as_sf(
    occ_with_region[unassigned_idx, ],
    coords = c("lon", "lat"), crs = 4326
  )

  # Find nearest country for each unassigned point
  nearest_idx <- st_nearest_feature(unassigned_sf, world)
  nearest_countries <- world$name[nearest_idx]

  # Re-assign regions
  occ_with_region$name[unassigned_idx] <- nearest_countries
  occ_with_region$region[unassigned_idx] <- case_when(
    nearest_countries %in% east_asia_countries ~ "East_Asia",
    nearest_countries %in% oceania_countries   ~ "Oceania",
    nearest_countries %in% north_america_countries ~ "North_America",
    TRUE ~ "Other"
  )

  sprintf("Re-assigned %d offshore/unmatched points using nearest-country", n_other)
} else {
  "All points matched to a country polygon — no re-assignment needed"
}
```

## 5. Geographically stratified thinning

Cap each region so no single region exceeds ~35% of total records. Regions already below the cap are left untouched; only over-represented regions are subsampled.

```{r}
#| label: stratified-thinning

set.seed(42)  # Reproducibility

total_records <- nrow(occ_with_region)
cap_pct <- 0.35
cap_n <- floor(total_records * cap_pct)

# Split by region, subsample those over the cap
occ_stratified <- occ_with_region %>%
  group_by(region) %>%
  group_modify(~ {
    if (nrow(.x) > cap_n) {
      slice_sample(.x, n = cap_n)
    } else {
      .x
    }
  }) %>%
  ungroup()

# Regional breakdown after stratified thinning
region_counts_after <- occ_stratified %>%
  count(region, name = "n_records") %>%
  mutate(pct = round(100 * n_records / sum(n_records), 1))

# Combined before/after comparison
region_comparison <- region_counts_before %>%
  rename(before_n = n_records, before_pct = pct) %>%
  left_join(
    region_counts_after %>% rename(after_n = n_records, after_pct = pct),
    by = "region"
  )

region_comparison
```

```{r}
#| label: thinning-summary

sprintf(
  "Stratified thinning: %d -> %d records (cap = %d per region, %.0f%%)",
  total_records, nrow(occ_stratified), cap_n, cap_pct * 100
)
```


## 6. Visualise the stratified dataset

```{r}
#| label: fig-stratified-map
#| fig-cap: "Occurrence records after geographically stratified thinning, coloured by region."
#| fig-width: 10
#| fig-height: 6

world_map <- ne_countries(scale = "medium", returnclass = "sf")

ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_point(
    data = occ_stratified,
    aes(x = lon, y = lat, colour = region),
    size = 1.5, alpha = 0.7
  ) +
  scale_colour_brewer(palette = "Set1", name = "Region") +
  coord_sf(xlim = c(90, 185), ylim = c(-50, 55)) +
  labs(
    title = "H. longicornis occurrences after stratified thinning",
    subtitle = sprintf("n = %d (capped at %d per region)", nrow(occ_stratified), cap_n),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(file.path(figures_dir, "01_stratified_occurrences_map.png"),
       width = 10, height = 6, dpi = 300)
```

```{r}
#| label: fig-region-bars
#| fig-cap: "Regional composition before and after stratified thinning."
#| fig-width: 8
#| fig-height: 4

region_comparison_long <- region_comparison %>%
  pivot_longer(
    cols = c(before_n, after_n),
    names_to = "stage",
    values_to = "n"
  ) %>%
  mutate(stage = ifelse(stage == "before_n", "Before thinning", "After thinning"))

ggplot(region_comparison_long, aes(x = region, y = n, fill = stage)) +
  geom_col(position = "dodge", width = 0.6) +
  geom_hline(yintercept = cap_n, linetype = "dashed", colour = "red", linewidth = 0.5) +
  annotate("text", x = 0.5, y = cap_n + 10, label = paste0("35% cap = ", cap_n),
           hjust = 0, colour = "red", size = 3.5) +
  scale_fill_manual(values = c("Before thinning" = "grey60", "After thinning" = "steelblue")) +
  labs(
    title = "Regional record counts before and after stratified thinning",
    x = "Region", y = "Number of records", fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(file.path(figures_dir, "01_region_comparison_bars.png"),
       width = 8, height = 4, dpi = 300)
```


## 7. Save outputs

```{r}
#| label: save-outputs

# Save the stratified occurrence data (with region assignments)
# Keep species, lon, lat, ID, region — drop the old climate columns
# (climate values will be re-extracted in Step 2 with the expanded variable set)
occ_output <- occ_stratified %>%
  select(species, lon, lat, ID, region)

write_csv(occ_output, file.path(processed_v2, "occurrences_stratified.csv"))

# Also save as an sf geopackage for spatial operations in later steps
occ_output_sf <- st_as_sf(occ_output, coords = c("lon", "lat"), crs = 4326)
st_write(occ_output_sf, file.path(processed_v2, "occurrences_stratified.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

sprintf("Saved %d records to processed_data_2/occurrences_stratified.csv", nrow(occ_output))
```


## Summary

```{r}
#| label: summary-table

tibble(
  Step = c(
    "v1 thinned dataset",
    "After removing NA coordinates",
    "After stratified thinning (35% cap)"
  ),
  Records = c(nrow(occ_raw), nrow(occ_clean), nrow(occ_stratified))
)
```


## Session Information

```{r}
#| label: session-info
sessionInfo()
```
