---
title: "Step 2: Environmental Data — Expanded Variable Set"
author: "Alexander W. Gofton"
date: today
format: gfm
---

## Overview

This script assembles an expanded set of environmental variables for SDM modeling,
going beyond the original 19 WorldClim bioclimatic variables to include
humidity and moisture metrics that are ecologically critical for *H. longicornis*.

**Data sources:**

1. **WorldClim 2.1** (existing) — 19 bioclimatic variables at 2.5 arcmin
2. **ENVIREM** — Climatic moisture index (CMI), aridity index, and PET, computed
   from WorldClim 2.1 using the `envirem` R package (Title & Bemmels 2018)
3. **CHELSA v2.1** — Mean annual vapor pressure deficit (VPD), downloaded at
   30 arcsec and aggregated to 2.5 arcmin

**Resolution justification:** 2.5 arcmin (~5 km at the equator) is used because
(a) it matches the spatial precision of occurrence records after 10 km thinning,
(b) it is the standard resolution for continental-scale *H. longicornis* SDMs
in the literature (Raghavan et al. 2019, Namgyal et al. 2020), and
(c) finer resolution would introduce microclimatic noise not captured by
occurrence record precision.

```{r}
#| label: setup
library(tidyverse)
library(terra)
library(sf)
library(geodata)
```

```{r}
#| label: paths

base_dir <- here::here()
processed_v1 <- file.path(base_dir, "processed_data")
processed_v2 <- file.path(base_dir, "processed_data_2")
figures_dir  <- file.path(base_dir, "figures_2")
temp_dir     <- file.path(base_dir, "temp")

# WorldClim data lives here from the v1 pipeline
worldclim_path <- file.path(base_dir, "data", "worldclim", "climate", "wc2.1_2.5m")

terraOptions(memfrac = 0.6, tempdir = temp_dir)
```


## 1. Load existing WorldClim 2.1 bioclimatic variables

These were downloaded in the v1 pipeline at 2.5 arcmin global extent.

```{r}
#| label: load-worldclim

# Build file paths for all 19 bioclim variables
bio_files <- paste0(worldclim_path, "/wc2.1_2.5m_bio_", 1:19, ".tif")

# Check which files exist
existing <- file.exists(bio_files)
if (!all(existing)) {
  warning(sprintf("Missing WorldClim files: %s", paste(bio_files[!existing], collapse = ", ")))
}

# Load as a multi-layer SpatRaster
bioclim_all <- rast(bio_files[existing])

# Rename layers to clean short names
names(bioclim_all) <- paste0("bio", 1:sum(existing))

sprintf("Loaded %d WorldClim bioclimatic layers", nlyr(bioclim_all))
sprintf("Resolution: %.4f degrees (~%.1f km)", res(bioclim_all)[1], res(bioclim_all)[1] * 111)
sprintf("Extent: %.1f to %.1f (lon), %.1f to %.1f (lat)",
        ext(bioclim_all)[1], ext(bioclim_all)[2],
        ext(bioclim_all)[3], ext(bioclim_all)[4])
```

### WorldClim unit notes

WorldClim 2.1 bioclimatic variables use the following units. Temperature variables
are stored in degrees Celsius (NOT multiplied by 10, unlike WorldClim v1.4).
Precipitation variables are in millimetres.

```{r}
#| label: worldclim-units

# Quick sanity check — BIO1 (annual mean temp) should be roughly -30 to +30 °C
bio1_range <- global(bioclim_all[["bio1"]], range, na.rm = TRUE)
sprintf("BIO1 range: %.1f to %.1f °C", bio1_range[1, 1], bio1_range[1, 2])
```


## 2. Compute ENVIREM variables from WorldClim 2.1

The ENVIREM dataset (Title & Bemmels 2018) provides ecologically meaningful
variables derived from temperature and precipitation. Pre-computed ENVIREM
rasters are based on WorldClim v1.4, so we compute them fresh from WorldClim 2.1
using the `envirem` R package for consistency.

We need monthly temperature min/max and monthly precipitation as inputs.

```{r}
#| label: download-monthly-worldclim

# Download monthly climate data if not already present
# These are needed as inputs for ENVIREM variable calculation
tmin_path <- file.path(base_dir, "data", "worldclim", "climate", "wc2.1_2.5m")
tmax_path <- tmin_path
prec_path <- tmin_path

# Check if monthly files already exist (from geodata download)
tmin_file_check <- file.path(tmin_path, "wc2.1_2.5m_tmin_01.tif")

if (!file.exists(tmin_file_check)) {
  message("Downloading monthly WorldClim data for ENVIREM computation...")
  # geodata::worldclim_global downloads to a cache; we load from there
  tmin_rast <- worldclim_global(var = "tmin", res = 2.5,
                                path = file.path(base_dir, "data", "worldclim"))
  tmax_rast <- worldclim_global(var = "tmax", res = 2.5,
                                path = file.path(base_dir, "data", "worldclim"))
  prec_rast <- worldclim_global(var = "prec", res = 2.5,
                                path = file.path(base_dir, "data", "worldclim"))
} else {
  # Load existing monthly files
  tmin_files <- list.files(tmin_path, pattern = "wc2.1_2.5m_tmin_\\d+\\.tif$", full.names = TRUE)
  tmax_files <- list.files(tmax_path, pattern = "wc2.1_2.5m_tmax_\\d+\\.tif$", full.names = TRUE)
  prec_files <- list.files(prec_path, pattern = "wc2.1_2.5m_prec_\\d+\\.tif$", full.names = TRUE)

  tmin_rast <- rast(sort(tmin_files))
  tmax_rast <- rast(sort(tmax_files))
  prec_rast <- rast(sort(prec_files))
}

sprintf("Monthly tmin layers: %d", nlyr(tmin_rast))
sprintf("Monthly tmax layers: %d", nlyr(tmax_rast))
sprintf("Monthly prec layers: %d", nlyr(prec_rast))
```

```{r}
#| label: compute-envirem
#| message: false

# Compute ENVIREM variables using the envirem R package
# If envirem is not installed, install it first
if (!requireNamespace("envirem", quietly = TRUE)) {
  install.packages("envirem")
}
library(envirem)

# envirem works with raster package objects, so we need to convert.
# However, if it supports terra objects (newer versions), we use those directly.
# The package expects: tmin (12 layers), tmax (12 layers), precip (12 layers)

# Generate a solar radiation raster (required for PET calculation)
# This uses the Hargreaves method based on latitude and extraterrestrial radiation
# envirem::ETsolradRasters expects a template raster to derive latitude
message("Computing monthly solar radiation (Hargreaves method)...")
solrad <- ETsolradRasters(tmin_rast, tmax_rast, tempScale = 1)

# Compute all ENVIREM variables
message("Computing ENVIREM variables...")
envirem_all <- layerCreation(
  tmin_rast, tmax_rast, prec_rast,
  sol674 = solrad,
  tempScale = 1  # WorldClim 2.1 is in °C, not °C * 10
)

sprintf("Computed %d ENVIREM variables:", nlyr(envirem_all))
names(envirem_all)
```

```{r}
#| label: select-envirem

# Extract the three ENVIREM variables we want:
# 1. Climatic Moisture Index (CMI) — ratio of precipitation to PET
# 2. Thornthwaite Aridity Index
# 3. Annual PET (potential evapotranspiration)

# ENVIREM variable names vary by package version — identify the right ones
envirem_names <- names(envirem_all)

# Match by partial name (case-insensitive)
cmi_idx     <- grep("climaticMoistureIndex|cmi", envirem_names, ignore.case = TRUE)
aridity_idx <- grep("aridityIndex|thornthwaite", envirem_names, ignore.case = TRUE)
pet_idx     <- grep("annualPET|PETann", envirem_names, ignore.case = TRUE)

# If exact matches fail, print available names for manual selection
if (length(cmi_idx) == 0 | length(aridity_idx) == 0 | length(pet_idx) == 0) {
  warning("Could not auto-match all ENVIREM layers. Available names:")
  print(envirem_names)
  message("Please adjust the variable selection indices manually.")
}

# Select and rename
envirem_selected <- c(
  envirem_all[[cmi_idx[1]]],
  envirem_all[[aridity_idx[1]]],
  envirem_all[[pet_idx[1]]]
)
names(envirem_selected) <- c("cmi", "aridity", "pet_annual")

sprintf("Selected ENVIREM variables: %s", paste(names(envirem_selected), collapse = ", "))
```


## 3. Download and process CHELSA vapor pressure deficit

CHELSA v2.1 provides monthly mean vapor pressure deficit (vpd) at 30 arcsec
resolution (~1 km). We download annual mean VPD and aggregate to 2.5 arcmin
to match our other variables.

```{r}
#| label: download-chelsa-vpd

# CHELSA v2.1 VPD is available as monthly files
# We download all 12 months and compute the annual mean
chelsa_dir <- file.path(base_dir, "data", "chelsa")
dir.create(chelsa_dir, showWarnings = FALSE, recursive = TRUE)

# CHELSA v2.1 VPD URL pattern (1981-2010 climatologies)
# Format: CHELSA_vpd_<month>_1981-2010_V.2.1.tif
chelsa_base_url <- "https://os.zhdk.cloud.switch.ch/envicloud/chelsa/chelsa_V2/GLOBAL/climatologies/1981-2010/vpd/"

vpd_files <- file.path(chelsa_dir, sprintf("CHELSA_vpd_%02d_1981-2010_V.2.1.tif", 1:12))
vpd_urls  <- paste0(chelsa_base_url, sprintf("CHELSA_vpd_%02d_1981-2010_V.2.1.tif", 1:12))

# Download if not already present
for (i in 1:12) {
  if (!file.exists(vpd_files[i])) {
    message(sprintf("Downloading CHELSA VPD month %02d...", i))
    tryCatch(
      download.file(vpd_urls[i], vpd_files[i], mode = "wb", quiet = TRUE),
      error = function(e) warning(sprintf("Failed to download month %d: %s", i, e$message))
    )
  }
}

# Check how many downloaded successfully
vpd_exists <- file.exists(vpd_files)
sprintf("CHELSA VPD files downloaded: %d/12", sum(vpd_exists))
```

```{r}
#| label: process-chelsa-vpd

if (sum(vpd_exists) == 12) {
  # Load all monthly VPD rasters
  vpd_monthly <- rast(vpd_files)

  # CHELSA VPD is stored as Pa (Pascals) — convert to kPa for interpretability
  # Scale factor: values are in Pa * 100 (i.e., stored as integer, divide by 100 for Pa)
  # Then divide by 1000 for kPa
  # Check CHELSA documentation for exact scaling — adjust if needed
  vpd_monthly <- vpd_monthly / 100  # Pa to hPa if needed; verify with data range

  # Compute annual mean VPD
  vpd_annual <- mean(vpd_monthly)
  names(vpd_annual) <- "vpd_annual"

  # Aggregate from 30 arcsec to 2.5 arcmin (factor of 5)
  # 2.5 arcmin / 30 arcsec = 5
  vpd_2.5arcmin <- aggregate(vpd_annual, fact = 5, fun = "mean", na.rm = TRUE)

  # Resample to match the exact grid of our WorldClim data
  vpd_resampled <- resample(vpd_2.5arcmin, bioclim_all[[1]], method = "bilinear")
  names(vpd_resampled) <- "vpd_annual"

  sprintf("CHELSA VPD aggregated: %.4f° resolution", res(vpd_resampled)[1])
  sprintf("VPD range (annual mean): %.2f to %.2f",
          global(vpd_resampled, min, na.rm = TRUE)[[1]],
          global(vpd_resampled, max, na.rm = TRUE)[[1]])

} else {
  warning("Not all CHELSA VPD files downloaded. VPD will be excluded from analysis.")
  vpd_resampled <- NULL
}
```


## 4. Resample ENVIREM to match WorldClim grid

ENVIREM was computed from the same WorldClim inputs so should already be on
the same grid, but we resample to be safe.

```{r}
#| label: resample-envirem

envirem_resampled <- resample(envirem_selected, bioclim_all[[1]], method = "bilinear")

# Verify alignment
sprintf("WorldClim resolution: %.6f", res(bioclim_all)[1])
sprintf("ENVIREM resolution:   %.6f", res(envirem_resampled)[1])
sprintf("Grids aligned: %s", compareGeom(bioclim_all, envirem_resampled, stopOnError = FALSE))
```


## 5. Stack all environmental layers

Combine WorldClim bioclimatic variables, ENVIREM, and CHELSA VPD into a
single multi-layer raster.

```{r}
#| label: stack-all

# Build the combined raster stack
if (!is.null(vpd_resampled)) {
  env_all <- c(bioclim_all, envirem_resampled, vpd_resampled)
} else {
  env_all <- c(bioclim_all, envirem_resampled)
}

sprintf("Combined environmental stack: %d layers", nlyr(env_all))
names(env_all)
```


## 6. Extract climate values at occurrence points

```{r}
#| label: extract-climate

# Load the stratified occurrence data from Step 1
occ <- read_csv(file.path(processed_v2, "occurrences_stratified.csv"),
                show_col_types = FALSE)

# Extract climate values at each occurrence point
occ_coords <- occ %>% select(lon, lat) %>% as.data.frame()
climate_at_occ <- extract(env_all, occ_coords)

# Combine with occurrence data
occ_with_climate <- bind_cols(occ, climate_at_occ %>% select(-ID))

# Check for NAs (points falling in ocean or outside raster extent)
n_complete <- sum(complete.cases(occ_with_climate %>% select(starts_with("bio"), cmi, aridity, pet_annual)))
sprintf("Records with complete climate data: %d / %d", n_complete, nrow(occ_with_climate))

# Remove records with incomplete climate data
occ_with_climate <- occ_with_climate %>%
  filter(complete.cases(across(c(starts_with("bio"), cmi, aridity, pet_annual))))

sprintf("After removing incomplete records: %d", nrow(occ_with_climate))
```


## 7. Quick sanity checks

Verify that climate values at occurrence points fall within biologically
plausible ranges for *H. longicornis*.

```{r}
#| label: sanity-checks

# Key biological thresholds from the literature:
# - Annual mean temp (BIO1): ~4-20°C at known locations (Heath 2016)
# - Annual precip (BIO12): typically >600mm, often >1000mm
# - Cold tolerance limit: BIO6 (min temp coldest month) > -15°C

checks <- occ_with_climate %>%
  summarise(
    bio1_min  = round(min(bio1, na.rm = TRUE), 1),
    bio1_max  = round(max(bio1, na.rm = TRUE), 1),
    bio1_mean = round(mean(bio1, na.rm = TRUE), 1),
    bio12_min  = round(min(bio12, na.rm = TRUE), 0),
    bio12_max  = round(max(bio12, na.rm = TRUE), 0),
    bio12_mean = round(mean(bio12, na.rm = TRUE), 0),
    bio6_min   = round(min(bio6, na.rm = TRUE), 1),
    pct_below_1000mm = round(100 * mean(bio12 < 1000, na.rm = TRUE), 1),
    cmi_range  = sprintf("%.2f to %.2f", min(cmi, na.rm = TRUE), max(cmi, na.rm = TRUE)),
    vpd_range  = if ("vpd_annual" %in% names(occ_with_climate))
                   sprintf("%.1f to %.1f", min(vpd_annual, na.rm = TRUE), max(vpd_annual, na.rm = TRUE))
                 else "N/A"
  )

# Print as a transposed table for readability
checks_t <- checks %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value")

checks_t
```


## 8. Save outputs

```{r}
#| label: save-outputs

# Save the full environmental raster stack (global extent)
writeRaster(env_all, file.path(processed_v2, "env_all_global.tif"),
            overwrite = TRUE)

# Save the Australia/NZ subset for faster loading in later steps
aus_nz_ext <- ext(110, 180, -48, -8)
env_aus_nz <- crop(env_all, aus_nz_ext)
writeRaster(env_aus_nz, file.path(processed_v2, "env_all_aus_nz.tif"),
            overwrite = TRUE)

# Save occurrence data with climate values
write_csv(occ_with_climate, file.path(processed_v2, "occurrences_with_climate.csv"))

# Save as RDS for faster loading in R
saveRDS(env_all, file.path(processed_v2, "env_all_global.rds"))
saveRDS(occ_with_climate, file.path(processed_v2, "occurrences_with_climate.rds"))

sprintf("Saved: %d-layer global stack, Aus/NZ crop, and %d occurrence records with climate",
        nlyr(env_all), nrow(occ_with_climate))
```

## 9. Variable overview figure

```{r}
#| label: fig-variable-overview
#| fig-cap: "Distribution of key environmental variables at H. longicornis occurrence points."
#| fig-width: 12
#| fig-height: 8

# Select a subset of key variables to plot
plot_vars <- c("bio1", "bio10", "bio11", "bio12", "bio15", "cmi", "pet_annual")
if ("vpd_annual" %in% names(occ_with_climate)) plot_vars <- c(plot_vars, "vpd_annual")

# Nice labels
var_labels <- c(
  bio1 = "BIO1: Annual Mean Temp (°C)",
  bio10 = "BIO10: Mean Temp Warmest Qtr (°C)",
  bio11 = "BIO11: Mean Temp Coldest Qtr (°C)",
  bio12 = "BIO12: Annual Precipitation (mm)",
  bio15 = "BIO15: Precipitation Seasonality",
  cmi = "CMI: Climatic Moisture Index",
  pet_annual = "PET: Annual Evapotranspiration",
  vpd_annual = "VPD: Vapor Pressure Deficit"
)

occ_with_climate %>%
  select(region, all_of(plot_vars)) %>%
  pivot_longer(-region, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = plot_vars, labels = var_labels[plot_vars])) %>%
  ggplot(aes(x = value, fill = region)) +
  geom_histogram(bins = 30, alpha = 0.7, position = "identity") +
  facet_wrap(~ variable, scales = "free", ncol = 2) +
  scale_fill_brewer(palette = "Set1", name = "Region") +
  labs(title = "Environmental conditions at occurrence points by region",
       x = "Value", y = "Count") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

ggsave(file.path(figures_dir, "02_variable_distributions_by_region.png"),
       width = 12, height = 8, dpi = 300)
```


## Summary

```{r}
#| label: summary

tibble(
  Source = c("WorldClim 2.1", "ENVIREM (computed)", "CHELSA v2.1", "TOTAL"),
  Variables = c(
    paste0("bio1-bio19 (", 19, ")"),
    "cmi, aridity, pet_annual (3)",
    ifelse(!is.null(vpd_resampled), "vpd_annual (1)", "none (download failed)"),
    as.character(nlyr(env_all))
  ),
  Resolution = c("2.5 arcmin", "2.5 arcmin", "2.5 arcmin (aggregated from 30 arcsec)", "2.5 arcmin"),
  Note = c(
    "Existing from v1 pipeline",
    "Freshly computed from WC 2.1 monthly data via envirem R package",
    "Downloaded climatologies 1981-2010",
    ""
  )
)
```


## Session Information

```{r}
#| label: session-info
sessionInfo()
```
