---
title: "Step 4: Accessible Area and Background Sampling"
author: "Alexander W. Gofton"
date: today
format:
  gfm:
    toc: true
    toc-depth: 3
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    results: hold
---

## Overview

This script defines the species' accessible area ("M") and generates
background points for MaxEnt modeling. Key changes from v1:

- **Accessible area defined as a 1000 km buffered convex hull** around all
  occurrence points, intersected with land polygons. This replaces the v1
  approach of sampling background from the entire global land surface.
- **Target-group bias correction** using Ixodidae (tick family) GBIF records
  as a proxy for sampling effort. Background points are weighted toward
  areas where ticks have actually been surveyed.
- **10,000 background points** sampled within M, with bias weighting.

These changes address the most critical issue identified in the v1 critique:
global background sampling was inflating AUC and distorting variable importance.

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(MASS)       # kde2d for bias surface
library(geodata)
```

```{r}
#| label: paths

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/"
processed_v2 <- file.path(base_dir, "processed_data_2")
figures_dir  <- file.path(base_dir, "figures_2")
outputs_dir  <- file.path(base_dir, "outputs_2")
data_dir     <- file.path(base_dir, "data")
```


## 1. Load occurrence data and environmental rasters

```{r}
#| label: load-data

# Stratified occurrence data from Step 1
occ <- read_csv(file.path(processed_v2, "occurrences_stratified.csv"),
                show_col_types = FALSE)

# Selected environmental raster stack from Step 3
# (only the final selected variables, global extent)

env_selected <- readRDS(file.path(processed_v2, "env_selected_global.rds"))
crs(env_selected) <- "EPSG:4326"  # Ensure consistent CRS
sprintf("Occurrence records: %d", nrow(occ))
sprintf("Environmental layers: %d (%s)", nlyr(env_selected),
        paste(names(env_selected), collapse = ", "))

# After loading
for (i in 1:nlyr(env_selected)) {
  crs(env_selected[[i]]) <- "EPSG:4326"
}
```


## 2. Define the accessible area (M)

Create a convex hull around all occurrence points, buffer by 1000 km,
and intersect with global land polygons to exclude ocean.

The 1000 km buffer was chosen because:
- It captures the range of dispersal pathways (livestock trade, migratory birds)
  within a biologically plausible distance from known populations.
- It includes sufficient environmental variation for MaxEnt to distinguish
  suitable from unsuitable habitat without being so large that the
  discrimination task becomes trivially easy (as in the v1 global approach).


```{r}
#| label: create-accessible-area

# Convert occurrence points to sf
occ_sf <- st_as_sf(occ, coords = c("lon", "lat"), crs = 4326)

# Create convex hull around all occurrence points
# We need to combine all points into one geometry first
occ_combined <- st_union(occ_sf)
convex_hull <- st_convex_hull(occ_sf)

# Buffer by 1000 km
hull_projected <- st_transform(convex_hull, crs = "+proj=moll")
buffer_1000km <- st_buffer(hull_projected, dist = 1000000)  # 1000 km in meters
buffer_wgs84 <- st_transform(buffer_1000km, crs = 4326)

# Make the buffer geometry valid before intersection
buffer_wgs84 <- st_make_valid(buffer_wgs84)

# Get global land polygons and intersect with buffer to exclude ocean
world_land <- ne_countries(scale = "medium", returnclass = "sf") |>
  st_union()  # Dissolve into a single polygon

world_land <- st_make_valid(world_land)

# Use st_intersection with prepared geometries
accessible_area <- st_intersection(buffer_wgs84, world_land)

# Calculate area of the accessible region
area_km2 <- as.numeric(st_area(accessible_area)) / 1e6
sprintf("Accessible area: %.0f km²", area_km2)
```

```{r}
#| label: fig-accessible-area
#| fig-cap: "Accessible area (M) defined as a 1000 km buffered convex hull around occurrence points, clipped to land."
#| fig-width: 12
#| fig-height: 6

world_map <- ne_countries(scale = "medium", returnclass = "sf")

ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_sf(data = accessible_area, fill = "lightblue", alpha = 0.3,
          colour = "blue", linewidth = 0.5) +
  geom_sf(data = occ_sf, colour = "red", size = 0.8, alpha = 0.6) +
  labs(
    title = "Accessible area (M) for H. longicornis",
    subtitle = sprintf("1000 km buffered convex hull — %.0f km²", area_km2)
  ) +
  theme_minimal(base_size = 12)

ggsave(file.path(figures_dir, "04_accessible_area_map.png"),
       width = 12, height = 6, dpi = 300)
```


## 3. Mask environmental rasters to accessible area

Crop and mask the environmental stack to the accessible area so that
background points are only sampled from within M.

```{r}
#| label: mask-env-to-M

# Convert accessible area sf to a terra SpatVector for masking
accessible_vect <- vect(accessible_area)

# Crop and mask the environmental rasters to the accessible area
env_M <- crop(env_selected, accessible_vect)
env_M <- mask(env_M, accessible_vect)

# Check how many valid (non-NA) cells exist within M
n_cells <- global(!is.na(env_M[[1]]), sum)[[1]]
sprintf("Valid raster cells within accessible area: %.0f", n_cells)
```


## 4. Target-group bias correction

Download Ixodidae (tick family) occurrence records from GBIF within the
accessible area. These records represent "where people have looked for ticks"
and serve as a proxy for sampling effort. We then create a kernel density
surface from these records and use it to weight background point sampling.

This ensures the model compares tick presence against what's available in
places people have actually surveyed, rather than random landscape.

```{r}
#| label: download-ixodidae
#| message: false

ixodidae_all <- read.csv(
  paste0(
    data_dir,
    "/Ixodidae_background_points/0023290-260208012135463.csv"
  ),
  sep = "\t"
) |>
  dplyr::filter(!is.na(decimalLongitude), !is.na(decimalLatitude)) |>
  dplyr::select(lon = decimalLongitude, lat = decimalLatitude) |>
  dplyr::distinct()
```

```{r}
#| label: simplify-ixodidae-records
# Simplifying Ixodidae records
# Split the accessible area into individual polygons
aa_parts <- st_cast(accessible_area, "POLYGON")

# Step 1: Thin to ~1 point per 0.1° grid cell (~10 km)
# This is fine because we only need a smooth KDE bias surface,
# not exact point locations.
ixodidae_thinned <- ixodidae_all %>%
  mutate(
    lon_bin = round(lon, 1),
    lat_bin = round(lat, 1)
  ) %>%
  distinct(lon_bin, lat_bin, .keep_all = TRUE) %>%
  dplyr::select(lon, lat)

sprintf("After spatial thinning (0.1°): %d of %d points",
        nrow(ixodidae_thinned), nrow(ixodidae_all))
```

```{r}
# Step 2: Pre-filter by the union of per-polygon bounding boxes
# (much tighter than a single global bbox)
keep <- rep(FALSE, nrow(ixodidae_thinned))

for (k in seq_len(nrow(aa_parts))) {
  bb <- st_bbox(aa_parts[k, ])
  in_bb <- ixodidae_thinned$lon >= bb["xmin"] &
           ixodidae_thinned$lon <= bb["xmax"] &
           ixodidae_thinned$lat >= bb["ymin"] &
           ixodidae_thinned$lat <= bb["ymax"]
  keep <- keep | in_bb
}

ixodidae_bbox <- ixodidae_thinned[keep, ]
sprintf("After per-polygon bbox filter: %d points", nrow(ixodidae_bbox))
```

```{r}
# Step 3: Now the precise intersection is fast
ixodidae_sf <- st_as_sf(ixodidae_bbox, coords = c("lon", "lat"), crs = 4326)
ixodidae_in_AA <- st_intersection(ixodidae_sf, accessible_area)

coords <- st_coordinates(ixodidae_in_AA)
ixodidae_points <- tibble(lon = coords[, 1], lat = coords[, 2])

sprintf("Final Ixodidae bias points within M: %d", nrow(ixodidae_points))

# Cache for future runs
write_csv(ixodidae_points, paste0(processed_v2, "/ixodidae_points.csv"))
```


```{r map-to-check-background_points}
# Create a map of Ixodidae occurrences
ggplot() +
  # Base map
  geom_sf(data = world_map, fill = "grey95", color = "grey70", linewidth = 0.2) +
  # Ixodidae points - convert regular coordinates to points
  geom_point(data = ixodidae_points,
             aes(x = lon, y = lat),
             color = "darkgreen", size = 0.3, alpha = 0.3) +
  # Add H. longicornis occurrence points for reference
  geom_sf(data = occ_sf, color = "red", size = 1, alpha = 0.8) +
  # Add labels and title
  labs(
    title = "Ixodidae occurrences within the accessible area",
    subtitle = sprintf("%d Ixodidae records (green) and %d H. longicornis records (red)",
                      nrow(ixodidae_points), nrow(occ_sf)),
    x = "Longitude",
    y = "Latitude"
  ) +
  # Apply minimal theme
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.2)
  )
```

### Create bias surface from Ixodidae records

```{r}
#| label: create-bias-surface

# Create a 2D kernel density estimate from Ixodidae records
# This produces a smooth surface representing sampling effort
# We'll rasterize this to match our environmental grid

# Use the extent of the environmental raster within M
ext_M <- ext(env_M)

# Create the KDE using MASS::kde2d
# n = number of grid points in each direction (match raster resolution roughly)
kde_result <- kde2d(
  x = ixodidae_points$lon,
  y = ixodidae_points$lat,
  n = 200,  # grid resolution for KDE
  lims = c(ext_M[1], ext_M[2], ext_M[3], ext_M[4])
)

# Convert KDE result to a SpatRaster
bias_rast <- rast(
  list(
    x = kde_result$x,
    y = kde_result$y,
    z = kde_result$z
  )
)
crs(bias_rast) <- "EPSG:4326"

# Resample to match the environmental grid exactly
bias_resampled <- resample(bias_rast, env_M[[1]], method = "bilinear")

# Mask to M (set ocean/outside-M cells to NA)
bias_resampled <- mask(bias_resampled, env_M[[1]])

# Add a small floor value so cells with zero bias density still have
# a non-zero chance of being selected (avoids excluding entire regions)
floor_val <- global(bias_resampled, quantile, probs = 0.01, na.rm = TRUE)[[1]]
bias_resampled <- max(bias_resampled, floor_val)

# Normalise to [0, 1] for use as sampling weights
bias_min <- global(bias_resampled, min, na.rm = TRUE)[[1]]
bias_max <- global(bias_resampled, max, na.rm = TRUE)[[1]]
bias_weights <- (bias_resampled - bias_min) / (bias_max - bias_min)

names(bias_weights) <- "bias_weight"

sprintf("Bias surface created: %.4f to %.4f (normalised)",
        global(bias_weights, min, na.rm = TRUE)[[1]],
        global(bias_weights, max, na.rm = TRUE)[[1]])
```

```{r}
#| label: fig-bias-surface
#| fig-cap: "Sampling bias surface derived from Ixodidae GBIF records (kernel density)."
#| fig-width: 12
#| fig-height: 6

# Convert bias surface to a data frame for ggplot
bias_df <- as.data.frame(bias_weights, xy = TRUE) |>
  filter(!is.na(bias_weight))

ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_raster(data = bias_df, aes(x = x, y = y, fill = bias_weight)) +
  scale_fill_viridis_c(name = "Sampling\nbias weight", option = "inferno") +
  geom_sf(data = occ_sf, colour = "cyan", size = 0.6, alpha = 0.7) +
  coord_sf(xlim = c(ext_M[1], ext_M[2]), ylim = c(ext_M[3], ext_M[4])) +
  labs(
    title = "Sampling effort bias surface (Ixodidae GBIF records)",
    subtitle = sprintf("Based on %d Ixodidae occurrence records within M",
                       nrow(ixodidae_points))
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

ggsave(file.path(figures_dir, "04_bias_surface.png"),
       width = 12, height = 6, dpi = 300)
```


## 5. Sample background points

Generate 10,000 random background points within the accessible area,
weighted by the bias surface so that well-surveyed areas receive
proportionally more background points.

```{r}
#| label: sample-background
set.seed(7990)

n_bg <- 10000

# --- Bias-weighted background sampling ---
# 1. Get all valid (non-NA) cell indices and their coordinates
valid_cells <- which(!is.na(values(env_M[[1]])))
cell_coords <- xyFromCell(env_M, valid_cells)

# 2. Extract bias weights for valid cells
bias_vals <- values(bias_weights)[valid_cells]
bias_vals[is.na(bias_vals) | bias_vals <= 0] <- 1e-6  # floor at near-zero

# 3. Sample cell indices weighted by bias surface
sampled_idx <- sample(
  seq_along(valid_cells),
  size = n_bg,
  replace = FALSE,
  prob = bias_vals
)

# 4. Extract environmental values at sampled cells
sampled_cells <- valid_cells[sampled_idx]
bg_values <- extract(env_M, sampled_cells)

bg_points <- as.data.frame(cell_coords[sampled_idx, ]) |>
  rename(lon = x, lat = y) |>
  bind_cols(bg_values |> dplyr::select(-any_of("ID"))) |>
  filter(complete.cases(pick(everything())))

sprintf("Background points generated: %d", nrow(bg_points))
```

```{r}
#| label: fig-background-points
#| fig-cap: "Background points (n = 10,000) sampled within the accessible area with bias correction."
#| fig-width: 12
#| fig-height: 6

ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_sf(data = accessible_area, fill = NA, colour = "blue",
          linewidth = 0.5, linetype = "dashed") +
  geom_point(data = bg_points, aes(x = lon, y = lat),
             colour = "grey40", size = 0.3, alpha = 0.3) +
  geom_sf(data = occ_sf, colour = "red", size = 1, alpha = 0.7) +
  coord_sf(xlim = c(ext_M[1], ext_M[2]), ylim = c(ext_M[3], ext_M[4])) +
  labs(
    title = "Occurrence and background points within accessible area",
    subtitle = sprintf("%d presence (red) + %d background (grey) points",
                       nrow(occ), nrow(bg_points)),
    x = "Longitude", y = "Latitude"
  ) +
  theme_minimal(base_size = 12)

ggsave(file.path(figures_dir, "04_occurrence_background_map.png"),
       width = 12, height = 6, dpi = 300)
```


## 6. Compare background climate with global climate

Verify that the restricted background captures the relevant environmental
gradients without spanning the entire global climate space.

```{r}
#| label: fig-climate-comparison
#| fig-cap: "Climate space comparison: presence points vs. background within M."
#| fig-width: 10
#| fig-height: 8

# Get climate values at occurrence points
occ_climate <- read_csv(
  file.path(processed_v2, "occurrences_selected_vars.csv"),
  show_col_types = FALSE
)

# bg_points already has environmental values from spatSample(..., values = TRUE)
bg_climate <- bg_points |>
  dplyr::select(-lon, -lat) |>
  filter(complete.cases(pick(everything())))

# Get matching variable names that exist in both datasets
selected_vars <- intersect(names(bg_climate), names(occ_climate))

# Subset occurrence climate data with the correct variables - use direct column selection
occ_climate_sub <- occ_climate[, selected_vars, drop = FALSE]

# Combine for plotting
bg_long <- bg_climate |>
  pivot_longer(cols = selected_vars, names_to = "variable", values_to = "value") |>
  mutate(source = "Background (M)")

occ_long <- occ_climate_sub |>
  pivot_longer(cols = selected_vars, names_to = "variable", values_to = "value") |>
  mutate(source = "Presence")

climate_comparison <- bind_rows(bg_long, occ_long)

ggplot(climate_comparison, aes(x = value, fill = source)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  scale_fill_manual(values = c("Background (M)" = "grey60", "Presence" = "red")) +
  labs(
    title = "Environmental space: presence vs. background within accessible area",
    x = "Value", y = "Density", fill = NULL
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```



## 7. Save outputs

```{r}
#| label: save-outputs

# Save the accessible area polygon
st_write(accessible_area,
         file.path(processed_v2, "accessible_area_M.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

# Save the bias surface raster
writeRaster(bias_weights,
            file.path(processed_v2, "bias_surface.tif"),
            overwrite = TRUE)

# Save the environmental stack cropped to M
writeRaster(env_M,
            file.path(processed_v2, "env_selected_M.tif"),
            overwrite = TRUE)
saveRDS(env_M, file.path(processed_v2, "env_selected_M.rds"))

# Save background points with climate values
write_csv(bg_points, file.path(processed_v2, "background_points.csv"))

# Save occurrence coordinates as a simple lon/lat dataframe
# (needed by ENMeval in Step 5)
occ_coords <- occ #|> select(lon, lat)
write_csv(occ_coords, file.path(processed_v2, "occurrence_coords.csv"))

# Save background coordinates
bg_coords <- bg_points #|> dplyr::select(lon, lat)
write_csv(bg_coords, file.path(processed_v2, "background_coords.csv"))

sprintf("Saved accessible area polygon, bias surface, %d background points",
        nrow(bg_points))
```


## Summary

```{r}
#| label: summary

tibble(
  Item = c(
    "Accessible area (M)",
    "Buffer distance",
    "Area of M",
    "Bias correction source",
    "Ixodidae records for bias",
    "Background points",
    "Occurrence points",
    "Environmental variables"
  ),
  Value = c(
    "Convex hull + buffer, clipped to land",
    "1000 km",
    sprintf("%.0f km²", sum(area_km2)),  # Sum the area values
    "Ixodidae GBIF records (target-group)",
    as.character(nrow(ixodidae_points)),
    as.character(nrow(bg_points)),
    as.character(nrow(occ)),
    paste(names(env_selected), collapse = ", ")
  )
)
```


## Session Information

```{r}
#| label: session-info
sessionInfo()
```
