---
title: "Step 7: Current Prediction for Australia/NZ"
author: "Alexander W. Gofton"
date: today
format: gfm
---

## Overview

Project the optimized MaxEnt model to Australia and New Zealand under
current climate conditions. Key improvements over v1:

- **MESS masking** — areas with novel climate conditions (MESS < 0) are
  masked on the binary map and cross-hatched on the continuous map.
- **Principled threshold** — binary suitable/unsuitable classification uses
  the Maximum Sensitivity + Specificity (MaxSS) threshold instead of 0.5.
- **Area calculations** — suitable habitat area computed from the binary map.

```{r}
#| label: setup
#| message: false

options(java.parameters = "-Xmx8g")

library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(predicts)
library(rJava)
```

```{r}
#| label: paths

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/"
processed_v2 <- file.path(base_dir, "processed_data_2")
figures_dir  <- file.path(base_dir, "figures_2")
outputs_dir  <- file.path(base_dir, "outputs_2")
```


## 1. Load model, rasters, and threshold

```{r}
#| label: load-data

# Selected MaxEnt model
selected_model <- readRDS(file.path(processed_v2, "selected_maxent_model.rds"))

# Environmental rasters for Australia/NZ (selected variables only)
env_aus_nz <- rast(file.path(processed_v2, "env_selected_aus_nz.tif"))

# MESS raster (computed in Step 6)
mess_aus_nz <- rast(file.path(processed_v2, "mess_aus_nz.tif"))

# Primary threshold (MaxSS, computed in Step 6)
primary_threshold <- as.numeric(readLines(
  file.path(processed_v2, "primary_threshold.txt")
))

sprintf("Loaded model, env rasters (%d layers), and threshold (%.4f)",
        nlyr(env_aus_nz), primary_threshold)
```


## 2. Generate continuous suitability prediction

```{r}
#| label: predict-current

# Project the model to Australia/NZ
pred_current <- predict(env_aus_nz, selected_model, type = "cloglog")
names(pred_current) <- "suitability"

# Summary statistics
suit_stats <- global(pred_current, fun = c("mean", "min", "max", "sd"),
                     na.rm = TRUE)

sprintf("Suitability range: %.4f to %.4f (mean = %.4f, SD = %.4f)",
        suit_stats$min, suit_stats$max, suit_stats$mean, suit_stats$sd)
```


## 3. Apply binary threshold and MESS masking

```{r}
#| label: binary-and-mess

# Binary prediction using MaxSS threshold
pred_binary <- pred_current >= primary_threshold
names(pred_binary) <- "suitable"

# MESS mask: identify cells with novel climate (MESS < 0)
mess_mask <- mess_aus_nz >= 0  # TRUE = analogous, FALSE = novel

# Binary prediction with MESS masking (novel areas set to NA)
pred_binary_masked <- mask(pred_binary, mess_mask, maskvalues = FALSE)
names(pred_binary_masked) <- "suitable_mess_masked"

# Count cells
cell_area_km2 <- prod(res(pred_binary)) * 111^2 * cos(mean(ext(pred_binary)[3:4]) * pi / 180)

n_suitable <- global(pred_binary, sum, na.rm = TRUE)[[1]]
n_suitable_masked <- global(pred_binary_masked, sum, na.rm = TRUE)[[1]]
n_novel <- global(!mess_mask, sum, na.rm = TRUE)[[1]]
n_total_land <- global(!is.na(pred_binary), sum)[[1]]

area_suitable <- n_suitable * cell_area_km2
area_suitable_masked <- n_suitable_masked * cell_area_km2
area_novel <- n_novel * cell_area_km2

sprintf("Suitable habitat (before MESS mask): %.0f km² (%.1f%% of land)",
        area_suitable, 100 * n_suitable / n_total_land)
sprintf("Suitable habitat (after MESS mask):  %.0f km² (%.1f%% of land)",
        area_suitable_masked, 100 * n_suitable_masked / n_total_land)
sprintf("Novel climate area (MESS < 0):       %.0f km² (%.1f%% of land)",
        area_novel, 100 * n_novel / n_total_land)
```


## 4. Maps

```{r}
#| label: fig-continuous-suitability
#| fig-cap: "Continuous habitat suitability for H. longicornis under current climate."
#| fig-width: 10
#| fig-height: 8

# Convert to data frame for ggplot
pred_df <- as.data.frame(pred_current, xy = TRUE) %>%
  filter(!is.na(suitability))

world_map <- ne_countries(scale = "medium", returnclass = "sf")

# Load occurrence points for overlay
occ <- read_csv(file.path(processed_v2, "occurrences_stratified.csv"),
                show_col_types = FALSE) %>%
  filter(region == "Oceania")  # Only Australian/NZ occurrences

ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_raster(data = pred_df, aes(x = x, y = y, fill = suitability)) +
  scale_fill_viridis_c(
    name = "Habitat\nsuitability",
    option = "magma",
    limits = c(0, 1)
  ) +
  geom_point(data = occ, aes(x = lon, y = lat),
             colour = "cyan", size = 1, alpha = 0.7, shape = 16) +
  coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
  labs(
    title = "H. longicornis habitat suitability — current climate",
    subtitle = sprintf("Cyan points = Australian/NZ occurrences | Threshold (MaxSS) = %.3f",
                       primary_threshold)
  ) +
  theme_minimal(base_size = 12)

ggsave(file.path(figures_dir, "07_current_suitability_continuous.png"),
       width = 10, height = 8, dpi = 300)
```

```{r}
#| label: fig-binary-suitability
#| fig-cap: "Binary suitable/unsuitable habitat (MaxSS threshold) with MESS masking."
#| fig-width: 10
#| fig-height: 8

# Create a three-class raster: suitable, unsuitable, novel (MESS < 0)
# 0 = unsuitable, 1 = suitable, 2 = novel climate (unreliable)
pred_classified <- pred_binary
pred_classified[!mess_mask & !is.na(pred_binary)] <- 2

class_df <- as.data.frame(pred_classified, xy = TRUE) %>%
  filter(!is.na(suitable)) %>%
  mutate(
    category = case_when(
      suitable == 2 ~ "Novel climate (MESS < 0)",
      suitable == 1 ~ "Suitable",
      suitable == 0 ~ "Unsuitable"
    ),
    category = factor(category,
                      levels = c("Suitable", "Unsuitable", "Novel climate (MESS < 0)"))
  )

ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_raster(data = class_df, aes(x = x, y = y, fill = category)) +
  scale_fill_manual(
    values = c(
      "Suitable" = "#2166AC",
      "Unsuitable" = "grey85",
      "Novel climate (MESS < 0)" = "#FFD700"
    ),
    name = NULL
  ) +
  geom_point(data = occ, aes(x = lon, y = lat),
             colour = "red", size = 1, alpha = 0.7, shape = 16) +
  coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
  labs(
    title = "H. longicornis — binary habitat suitability (current climate)",
    subtitle = sprintf(
      "Suitable: %.0f km² | Novel climate: %.0f km² | Threshold: %.3f (MaxSS)",
      area_suitable, area_novel, primary_threshold
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

ggsave(file.path(figures_dir, "07_current_suitability_binary.png"),
       width = 10, height = 8, dpi = 300)
```


## 5. Save outputs

```{r}
#| label: save-outputs

# Save rasters
writeRaster(pred_current,
            file.path(processed_v2, "prediction_current_continuous.tif"),
            overwrite = TRUE)
writeRaster(pred_binary,
            file.path(processed_v2, "prediction_current_binary.tif"),
            overwrite = TRUE)
writeRaster(pred_binary_masked,
            file.path(processed_v2, "prediction_current_binary_mess_masked.tif"),
            overwrite = TRUE)

# Save area statistics
area_stats <- tibble(
  Measure = c(
    "Suitable habitat (before MESS mask)",
    "Suitable habitat (after MESS mask)",
    "Novel climate area (MESS < 0)",
    "Total land area (Aus/NZ extent)",
    "Threshold (MaxSS)"
  ),
  Value = c(
    sprintf("%.0f km²", area_suitable),
    sprintf("%.0f km²", area_suitable_masked),
    sprintf("%.0f km²", area_novel),
    sprintf("%.0f km²", n_total_land * cell_area_km2),
    sprintf("%.4f", primary_threshold)
  )
)

write_csv(area_stats, file.path(outputs_dir, "current_habitat_area_v2.csv"))

sprintf("Current prediction outputs saved")
```


## Summary

```{r}
#| label: summary

area_stats
```


## Session Information

```{r}
#| label: session-info
sessionInfo()
```
