---
title: "Step 8: Future Climate Data"
author: "Alexander W. Gofton"
date: today
format: gfm
---

## Overview

Download and prepare future climate data for 5 GCMs, 2 SSPs, and 2 time
periods (20 total future climate stacks). Key changes from v1:

- **5 GCMs** (was 3): MIROC6, MPI-ESM1-2-HR, UKESM1-0-LL + ACCESS-CM2
  and CNRM-CM6-1
- **Match all selected variables** including ENVIREM and CHELSA-derived
  layers (using delta-change method where future versions aren't directly
  available)
- **Explicit temperature unit documentation**

**Scenario matrix:** 5 GCMs × 2 SSPs × 2 time periods = 20 stacks

**Note on ENVIREM/CHELSA future variables:** Direct future projections for
ENVIREM and CHELSA are not available for all GCMs. We use a delta-change
approach: compute the proportional change in the underlying WorldClim
variables between current and future, and apply that change to the current
ENVIREM/CHELSA values. This is a standard approach in SDM studies (e.g.,
Raghavan et al. 2019).

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(terra)
library(geodata)
```

```{r}
#| label: paths

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/"
processed_v2 <- file.path(base_dir, "processed_data_2")
figures_dir  <- file.path(base_dir, "figures_2")
outputs_dir  <- file.path(base_dir, "outputs_2")
data_dir     <- file.path(base_dir, "data")
temp_dir     <- file.path(base_dir, "temp")

terraOptions(memfrac = 0.6, tempdir = temp_dir)
```


## 1. Define scenario matrix

```{r}
#| label: define-scenarios

# GCMs — 5 models spanning a range of climate sensitivities
gcms <- c(
  "MIROC6",          # Japanese model, moderate-high sensitivity
  "MPI-ESM1-2-HR",   # German model, moderate sensitivity
  "UKESM1-0-LL",     # UK model, high sensitivity
  "ACCESS-CM2",      # Australian model, good regional performance for Oceania
  "CNRM-CM6-1"       # French model, moderate sensitivity
)

# SSPs (Shared Socioeconomic Pathways)
ssps <- c("245", "585")
# SSP2-4.5: "middle of the road" — moderate emissions
# SSP5-8.5: "fossil-fuelled development" — high emissions

# Time periods
periods <- c("2041-2060", "2081-2100")

# Full scenario grid
scenarios <- expand_grid(
  gcm = gcms,
  ssp = ssps,
  period = periods
)

sprintf("Total future scenarios: %d", nrow(scenarios))
scenarios
```


## 2. Load current climate data

We need the current WorldClim bioclimatic variables both for the final
selected variables and for computing the delta-change ratios for
ENVIREM/CHELSA layers.

```{r}
#| label: load-current

# Current environmental stack (selected variables, Aus/NZ)
env_current <- rast(file.path(processed_v2, "env_selected_aus_nz.tif"))
selected_vars <- names(env_current)

sprintf("Selected variables for projection: %s",
        paste(selected_vars, collapse = ", "))

# Identify which variables come from each source
worldclim_vars <- selected_vars[grepl("^bio", selected_vars)]
envirem_vars   <- selected_vars[grepl("(cmi_idx|aridity_idx|pet_idx)", selected_vars)]
chelsa_vars    <- selected_vars[grepl("vpd_annual", selected_vars)]

sprintf("WorldClim: %s", paste(worldclim_vars, collapse = ", "))
sprintf("ENVIREM:   %s", paste(envirem_vars, collapse = ", "))
sprintf("CHELSA:    %s", paste(chelsa_vars, collapse = ", "))

# Load current WorldClim full global stack (for delta-change calculations)
# We need current bio12 (annual precip) and bio1 (mean temp) as baselines
# for scaling ENVIREM/CHELSA to future conditions
worldclim_path <- file.path(data_dir, "worldclim", "climate", "wc2.1_2.5m")
```


## 3. Download future WorldClim bioclimatic variables

Download future bioclimatic variables for all 20 scenarios. The geodata
package handles CMIP6 data from WorldClim.

```{r}
#| label: download-future-worldclim
#| message: false

# Directory for future climate data
future_dir <- file.path(data_dir, "future_climate")
dir.create(future_dir, showWarnings = FALSE, recursive = TRUE)

# Aus/NZ extent for cropping
aus_nz_ext <- ext(110, 180, -48, -8)

# Download and process each scenario
for (i in seq_len(nrow(scenarios))) {
  gcm    <- scenarios$gcm[i]
  ssp    <- scenarios$ssp[i]
  period <- scenarios$period[i]

  scenario_id <- sprintf("%s_ssp%s_%s", gcm, ssp, period)
  cache_file  <- file.path(future_dir, paste0(scenario_id, "_bioclim.tif"))

  if (file.exists(cache_file)) {
    message(sprintf("[%d/%d] %s — cached", i, nrow(scenarios), scenario_id))
    next
  }

  message(sprintf("[%d/%d] Downloading %s...", i, nrow(scenarios), scenario_id))

  tryCatch({
    # Download CMIP6 bioclimatic variables via geodata
    future_bio <- cmip6_world(
      model = gcm,
      ssp = ssp,
      time = period,
      var = "bioc",
      res = 2.5,
      path = file.path(data_dir, "worldclim")
    )

    # Rename layers to match our convention (bio1, bio2, ..., bio19)
    names(future_bio) <- paste0("bio", 1:nlyr(future_bio))

    # Crop to Aus/NZ
    future_bio_aus <- crop(future_bio, aus_nz_ext)

    # Save the cropped bioclim stack
    writeRaster(future_bio_aus, cache_file, overwrite = TRUE)

    message(sprintf("  Saved: %s (%d layers)", scenario_id, nlyr(future_bio_aus)))

  }, error = function(e) {
    message(sprintf("  ERROR downloading %s: %s", scenario_id, e$message))
  })
}
```


## 4. Compute future ENVIREM/CHELSA variables using delta-change method

For ENVIREM (CMI, aridity, PET) and CHELSA (VPD), we don't have direct
future projections from each GCM. Instead, we compute relative changes
in the underlying climate variables and apply them to the current
ENVIREM/CHELSA values.

**Delta-change logic:**

- **PET** scales primarily with temperature: `PET_future = PET_current × (bio1_future / bio1_current)`
- **CMI** (moisture index) scales with precipitation relative to PET:
  `CMI_future = CMI_current × (bio12_future / bio12_current) / (bio1_future / bio1_current)`
  (simplified: more precip → higher CMI, more temp → lower CMI)
- **Aridity** is inversely related to moisture: apply inverse of CMI change
- **VPD** scales roughly with temperature (exponential relationship):
  `VPD_future = VPD_current × exp(0.06 × (bio1_future - bio1_current))`
  (Clausius-Clapeyron: ~6% increase per 1°C warming)

```{r}
#| label: compute-delta-change

# Load current WorldClim bio1 and bio12 for Aus/NZ (for delta ratios)
bio1_current  <- rast(file.path(worldclim_path, "wc2.1_2.5m_bio_1.tif"))
bio12_current <- rast(file.path(worldclim_path, "wc2.1_2.5m_bio_12.tif"))
bio1_current  <- crop(bio1_current, aus_nz_ext)
bio12_current <- crop(bio12_current, aus_nz_ext)

# Load current ENVIREM/CHELSA layers for Aus/NZ
# These are already in the selected env stack
envirem_chelsa_current <- env_current[[c(envirem_vars, chelsa_vars)]]

# Process each scenario
for (i in seq_len(nrow(scenarios))) {
  gcm    <- scenarios$gcm[i]
  ssp    <- scenarios$ssp[i]
  period <- scenarios$period[i]

  scenario_id <- sprintf("%s_ssp%s_%s", gcm, ssp, period)
  output_file <- file.path(future_dir, paste0(scenario_id, "_all_selected.tif"))

  if (file.exists(output_file)) {
    message(sprintf("[%d/%d] %s — final stack cached", i, nrow(scenarios), scenario_id))
    next
  }

  # Load the future WorldClim bioclimatic variables
  bioclim_file <- file.path(future_dir, paste0(scenario_id, "_bioclim.tif"))
  if (!file.exists(bioclim_file)) {
    message(sprintf("  Skipping %s — bioclim file missing", scenario_id))
    next
  }

  future_bio <- rast(bioclim_file)

  # Extract selected WorldClim variables
  future_wc <- future_bio[[worldclim_vars]]

  # --- Delta-change for ENVIREM variables ---
  bio1_future  <- future_bio[["bio1"]]
  bio12_future <- future_bio[["bio12"]]

  # Temperature ratio (avoid division by zero with a small floor)
  temp_ratio <- bio1_future / max(bio1_current, 0.1)
  # Precipitation ratio
  precip_ratio <- bio12_future / max(bio12_current, 1)

  future_envirem_chelsa <- list()

  if ("pet_idx" %in% selected_vars) {
    # PET scales with temperature
    future_envirem_chelsa[["pet_idx"]] <-
      envirem_chelsa_current[["pet_idx"]] * abs(temp_ratio)
  }

  if ("cmi_idx" %in% selected_vars) {
    # CMI: moisture balance scales with precip/PET ratio
    future_envirem_chelsa[["cmi_idx"]] <-
      envirem_chelsa_current[["cmi_idx"]] * precip_ratio / max(abs(temp_ratio), 0.1)
  }

  if ("aridity_idx" %in% selected_vars) {
    # Aridity inversely related to moisture
    future_envirem_chelsa[["aridity_idx"]] <-
      envirem_chelsa_current[["aridity_idx"]] * abs(temp_ratio) / max(precip_ratio, 0.1)
  }

  if ("vpd_annual" %in% selected_vars) {
    # VPD follows Clausius-Clapeyron (~6% per °C)
    temp_change <- bio1_future - bio1_current
    future_envirem_chelsa[["vpd_annual"]] <-
      envirem_chelsa_current[["vpd_annual"]] * exp(0.06 * temp_change)
  }

  # Stack all future layers in the correct order
  future_stack <- future_wc

  for (var_name in c(envirem_vars, chelsa_vars)) {
    if (var_name %in% names(future_envirem_chelsa)) {
      future_stack <- c(future_stack, future_envirem_chelsa[[var_name]])
    }
  }

  # Verify layer names match selected variables
  names(future_stack) <- selected_vars

  # Ensure alignment with current env raster
  if (!compareGeom(future_stack, env_current, stopOnError = FALSE)) {
    future_stack <- resample(future_stack, env_current, method = "bilinear")
  }

  # Save
  writeRaster(future_stack, output_file, overwrite = TRUE)
  message(sprintf("[%d/%d] %s — saved (%d layers)",
                  i, nrow(scenarios), scenario_id, nlyr(future_stack)))
}
```


## 5. Verify all scenarios downloaded successfully

```{r}
#| label: verify-downloads

# Check which scenarios have complete data
download_status <- scenarios %>%
  mutate(
    scenario_id = sprintf("%s_ssp%s_%s", gcm, ssp, period),
    bioclim_file = file.path(future_dir, paste0(scenario_id, "_bioclim.tif")),
    final_file = file.path(future_dir, paste0(scenario_id, "_all_selected.tif")),
    bioclim_exists = file.exists(bioclim_file),
    final_exists = file.exists(final_file)
  )

sprintf("Scenarios with bioclim data: %d / %d",
        sum(download_status$bioclim_exists), nrow(download_status))
sprintf("Scenarios with complete stacks: %d / %d",
        sum(download_status$final_exists), nrow(download_status))

# Report any missing scenarios
missing <- download_status %>% filter(!final_exists)
if (nrow(missing) > 0) {
  message("WARNING: Missing scenario stacks:")
  print(missing %>% select(gcm, ssp, period))
}
```


## 6. Quick comparison of current vs. future climate

Verify that the future climate data shows expected warming and
precipitation changes relative to current conditions.

```{r}
#| label: fig-climate-comparison
#| fig-cap: "Distribution of BIO1 (annual mean temperature) across scenarios for Australia/NZ."
#| fig-width: 10
#| fig-height: 6

# Sample the current and future bio1 values across Aus/NZ
set.seed(99)
sample_cells <- spatSample(env_current[["bio1"]], size = 5000,
                           method = "random", na.rm = TRUE, xy = TRUE)
current_bio1 <- sample_cells$bio1

# Sample future bio1 for each scenario
future_bio1_data <- list()
future_bio1_data[["Current"]] <- tibble(bio1 = current_bio1, scenario = "Current")

for (i in seq_len(nrow(scenarios))) {
  scenario_id <- sprintf("%s_ssp%s_%s", scenarios$gcm[i], scenarios$ssp[i],
                         scenarios$period[i])
  final_file <- file.path(future_dir, paste0(scenario_id, "_all_selected.tif"))

  if (file.exists(final_file)) {
    future_stack <- rast(final_file)
    future_vals <- extract(future_stack[["bio1"]],
                           sample_cells[, c("x", "y")])$bio1
    future_bio1_data[[scenario_id]] <- tibble(
      bio1 = future_vals,
      scenario = sprintf("%s SSP%s %s",
                         scenarios$gcm[i], scenarios$ssp[i], scenarios$period[i])
    )
  }
}

bio1_comparison <- bind_rows(future_bio1_data)

ggplot(bio1_comparison, aes(x = bio1, fill = scenario)) +
  geom_density(alpha = 0.3) +
  labs(
    title = "Annual mean temperature (BIO1) distribution — current vs. future",
    x = "BIO1 (°C)", y = "Density", fill = "Scenario"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "right", legend.text = element_text(size = 7))

ggsave(file.path(figures_dir, "08_future_climate_comparison.png"),
       width = 10, height = 6, dpi = 300)
```


## 7. Document temperature units

**WorldClim 2.1 temperature units:**

- Temperature variables (BIO1–BIO11) are in **degrees Celsius** (not ×10).
- This applies to both current and future (CMIP6) data from WorldClim 2.1.
- WorldClim v1.4 stored temperature × 10 (as integers), but v2.1 uses
  floating point in actual °C.
- Precipitation variables (BIO12–BIO19) are in **millimetres**.

This is important for interpreting response curves and setting any manual
thresholds. A BIO1 value of 15 means 15°C (not 1.5°C).


## 8. Save scenario metadata

```{r}
#| label: save-metadata

# Save the scenario grid for reference in Step 9
write_csv(scenarios, file.path(processed_v2, "future_scenarios.csv"))

# Save the file paths for each scenario
scenario_paths <- download_status %>%
  select(gcm, ssp, period, scenario_id, final_file, final_exists) %>%
  filter(final_exists)
write_csv(scenario_paths, file.path(processed_v2, "future_scenario_paths.csv"))

sprintf("Saved metadata for %d future climate scenarios", nrow(scenario_paths))
```


## Summary

```{r}
#| label: summary

tibble(
  Item = c(
    "GCMs",
    "SSPs",
    "Time periods",
    "Total scenarios",
    "Successfully prepared",
    "WorldClim variables",
    "Delta-change variables",
    "Resolution"
  ),
  Details = c(
    paste(gcms, collapse = ", "),
    "SSP2-4.5, SSP5-8.5",
    "2041-2060, 2081-2100",
    as.character(nrow(scenarios)),
    as.character(sum(download_status$final_exists)),
    paste(worldclim_vars, collapse = ", "),
    paste(c(envirem_vars, chelsa_vars), collapse = ", "),
    "2.5 arcmin (~5 km)"
  )
)
```


## Session Information

```{r}
#| label: session-info
sessionInfo()
```
