---
title: "Step 10: Visualization and Results Summary"
author: "Alexander W. Gofton"
date: today
format: gfm
---

## Overview

Final visualization and results summary. This step produces publication-quality
figures and tables with key improvements from v1:

- **Range shift framing** — maps show gain, loss, and stable areas (not just
  net expansion)
- **Uncertainty prominence** — MESS extrapolation, GCM variability, and
  validation caveats presented upfront
- **Per-GCM results** alongside ensemble statistics
- **Priority surveillance map** — areas with high-confidence future suitability
  (suitable in ≥4 of 5 GCMs)

```{r}
#| label: setup
#| message: false

library(tidyverse)
library(terra)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(patchwork)
```

```{r}
#| label: paths

base_dir <- "~/Library/CloudStorage/OneDrive-CSIRO/OneDrive - Docs/01_Projects/Hlongicornis_SDM/"
processed_v2 <- file.path(base_dir, "processed_data_2")
figures_dir  <- file.path(base_dir, "figures_2")
outputs_dir  <- file.path(base_dir, "outputs_2")
```


## 1. Load all results

```{r}
#| label: load-results

# Current prediction
pred_current <- rast(file.path(processed_v2, "prediction_current_continuous.tif"))
pred_binary  <- rast(file.path(processed_v2, "prediction_current_binary.tif"))
mess_current <- rast(file.path(processed_v2, "mess_aus_nz.tif"))

# Threshold
primary_threshold <- as.numeric(readLines(
  file.path(processed_v2, "primary_threshold.txt")
))

# Model evaluation results
metrics      <- read_csv(file.path(outputs_dir, "model_metrics_v2.csv"),
                         show_col_types = FALSE)
comparison   <- read_csv(file.path(outputs_dir, "v1_v2_comparison.csv"),
                         show_col_types = FALSE)
validation   <- read_csv(file.path(outputs_dir, "biological_validation_v2.csv"),
                         show_col_types = FALSE)
loro_results <- read_csv(file.path(outputs_dir, "loro_cv_results_v2.csv"),
                         show_col_types = FALSE)
niche_overlap <- read_csv(file.path(outputs_dir, "niche_overlap_v2.csv"),
                          show_col_types = FALSE)

# Future projection results
shift_table <- read_csv(file.path(outputs_dir, "range_shift_summary_v2.csv"),
                        show_col_types = FALSE)
gcm_table   <- read_csv(file.path(outputs_dir, "per_gcm_results_v2.csv"),
                         show_col_types = FALSE)
uncertainty <- read_csv(file.path(outputs_dir, "ensemble_uncertainty_v2.csv"),
                        show_col_types = FALSE)

# Variable importance
var_importance <- read_csv(file.path(outputs_dir, "variable_importance_v2.csv"),
                           show_col_types = FALSE)

# Occurrence data (for overlay)
occ <- read_csv(file.path(processed_v2, "occurrences_stratified.csv"),
                show_col_types = FALSE)
occ_aus <- occ %>% filter(region == "Oceania")

# Scenario metadata
scenarios <- read_csv(file.path(processed_v2, "future_scenarios.csv"),
                      show_col_types = FALSE)

# World map for plotting
world_map <- ne_countries(scale = "medium", returnclass = "sf")
aus_nz <- ne_countries(scale = "medium", returnclass = "sf") %>%
  filter(iso_a3 %in% c("AUS", "NZL"))
```


## 2. Key caveats and uncertainty summary

This section is placed first deliberately — any interpretation of the
results below should be read in context of these limitations.

```{r}
#| label: caveats-summary

# Compile all uncertainty/limitation information
caveats <- tibble(
  Category = c(
    "Extrapolation (current)",
    "GCM agreement",
    "Biological validation — east coast",
    "Biological validation — deserts",
    "Niche overlap (native vs invaded)",
    "Delta-change approximation",
    "No dispersal constraints"
  ),
  Detail = c(
    sprintf("%.1f%% of Aus/NZ cells have novel climate (MESS < 0)",
            100 * global(mess_current < 0, sum, na.rm = TRUE)[[1]] /
              global(!is.na(mess_current), sum)[[1]]),
    sprintf("GCM range in net change: %s",
            paste(uncertainty$range_pct_change, collapse = " | ")),
    sprintf("%s — %s (v1 was 0.157 = FAIL)",
            validation$Value[1], validation$Status[1]),
    sprintf("%s — %s",
            validation$Value[2], validation$Status[2]),
    sprintf("Schoener's D = %.3f, Warren's I = %.3f",
            niche_overlap$Value[1], niche_overlap$Value[2]),
    "ENVIREM/CHELSA future values computed via delta-change method (not direct GCM output)",
    "Assumes no dispersal limitation (justified: open pastoral landscape)"
  )
)

caveats
```


## 3. Publication Figure 1: Current habitat suitability

```{r}
#| label: fig-pub-current
#| fig-cap: "Figure 1. Current habitat suitability for H. longicornis in Australia and New Zealand."
#| fig-width: 12
#| fig-height: 10

# Continuous suitability map
pred_df <- as.data.frame(pred_current, xy = TRUE) %>%
  filter(!is.na(suitability))

mess_df <- as.data.frame(mess_current, xy = TRUE) %>%
  filter(!is.na(mess), mess < 0)

p_current <- ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_raster(data = pred_df, aes(x = x, y = y, fill = suitability)) +
  # Overlay MESS < 0 as semi-transparent hatching
  geom_raster(data = mess_df, aes(x = x, y = y), fill = "grey50", alpha = 0.3) +
  scale_fill_viridis_c(
    name = "Habitat\nsuitability",
    option = "magma",
    limits = c(0, 1)
  ) +
  geom_point(data = occ_aus, aes(x = lon, y = lat),
             colour = "cyan", size = 1.2, alpha = 0.8, shape = 16) +
  coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
  labs(
    title = "A) Continuous habitat suitability (current climate)",
    subtitle = "Grey shading = novel climate (MESS < 0) | Cyan = occurrence records"
  ) +
  theme_minimal(base_size = 11)

# Binary suitability map (with MESS classes)
binary_df <- as.data.frame(pred_binary, xy = TRUE) %>%
  filter(!is.na(suitable))

mess_mask_df <- as.data.frame(mess_current >= 0, xy = TRUE) %>%
  filter(!is.na(mess))

combined_df <- binary_df %>%
  left_join(
    as.data.frame(mess_current, xy = TRUE) %>% rename(mess_val = mess),
    by = c("x", "y")
  ) %>%
  mutate(
    category = case_when(
      mess_val < 0 ~ "Novel climate",
      suitable == 1 ~ "Suitable",
      TRUE ~ "Unsuitable"
    )
  )

p_binary <- ggplot() +
  geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
  geom_raster(data = combined_df, aes(x = x, y = y, fill = category)) +
  scale_fill_manual(
    values = c(
      "Suitable" = "#2166AC",
      "Unsuitable" = "grey90",
      "Novel climate" = "#FFD700"
    ),
    name = NULL
  ) +
  geom_point(data = occ_aus, aes(x = lon, y = lat),
             colour = "red", size = 1, alpha = 0.8, shape = 16) +
  coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
  labs(
    title = sprintf("B) Binary suitability (threshold = %.3f, MaxSS)", primary_threshold),
    subtitle = "Red = occurrence records"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

p_current / p_binary

ggsave(file.path(figures_dir, "10_fig1_current_suitability.png"),
       width = 12, height = 10, dpi = 300)
```


## 4. Publication Figure 2: Range shift maps

```{r}
#| label: fig-pub-range-shift
#| fig-cap: "Figure 2. Projected range shifts under four future climate scenarios (ensemble median)."
#| fig-width: 12
#| fig-height: 10

ssp_periods <- expand_grid(
  ssp = c("245", "585"),
  period = c("2041-2060", "2081-2100")
)

shift_plots <- list()

for (j in seq_len(nrow(ssp_periods))) {
  group_id <- sprintf("ssp%s_%s", ssp_periods$ssp[j], ssp_periods$period[j])
  shift_file <- file.path(processed_v2, paste0("range_shift_", group_id, ".tif"))

  if (!file.exists(shift_file)) next

  shift_rast <- rast(shift_file)
  shift_df <- as.data.frame(shift_rast, xy = TRUE) %>%
    filter(!is.na(shift_class)) %>%
    mutate(
      category = case_when(
        shift_class == 3 ~ "Stable suitable",
        shift_class == 1 ~ "Gain",
        shift_class == 2 ~ "Loss",
        shift_class == 0 ~ "Stable unsuitable"
      ),
      category = factor(category,
                        levels = c("Gain", "Stable suitable", "Loss", "Stable unsuitable"))
    )

  # Get stats for this scenario
  stats <- shift_table %>% filter(scenario == group_id)

  nice_label <- sprintf("SSP%s %s",
                         ssp_periods$ssp[j],
                         ssp_periods$period[j])

  p <- ggplot() +
    geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
    geom_raster(data = shift_df %>% filter(category != "Stable unsuitable"),
                aes(x = x, y = y, fill = category)) +
    scale_fill_manual(
      values = c(
        "Gain" = "#2ca02c",
        "Stable suitable" = "#1f77b4",
        "Loss" = "#d62728"
      ),
      name = NULL,
      drop = FALSE
    ) +
    coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
    labs(
      title = nice_label,
      subtitle = sprintf("Net: %+.0f km² (%+.1f%%)",
                         stats$net_change_km2, stats$pct_change)
    ) +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")

  shift_plots[[group_id]] <- p
}

if (length(shift_plots) == 4) {
  wrap_plots(shift_plots, ncol = 2) +
    plot_annotation(
      title = "Range shift projections for H. longicornis (ensemble median, 5 GCMs)",
      subtitle = "Green = habitat gain | Blue = stable suitable | Red = habitat loss"
    )

  ggsave(file.path(figures_dir, "10_fig2_range_shift_maps.png"),
         width = 12, height = 10, dpi = 300)
}
```


## 5. Publication Figure 3: GCM uncertainty

```{r}
#| label: fig-pub-uncertainty
#| fig-cap: "Figure 3. Inter-GCM variability in projected habitat change."
#| fig-width: 10
#| fig-height: 8

# Panel A: Box plot of net change across GCMs
p_box <- gcm_table %>%
  mutate(
    scenario_label = sprintf("SSP%s\n%s", ssp, period)
  ) %>%
  ggplot(aes(x = scenario_label, y = pct_change)) +
  geom_boxplot(fill = "lightblue", outlier.shape = NA) +
  geom_jitter(aes(colour = gcm), width = 0.15, size = 2.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(
    title = "A) Inter-GCM variability in net habitat change",
    x = NULL, y = "Net change (%)", colour = "GCM"
  ) +
  theme_minimal(base_size = 11)

# Panel B: Extrapolation by scenario
p_extrap <- gcm_table %>%
  mutate(
    scenario_label = sprintf("SSP%s\n%s", ssp, period)
  ) %>%
  ggplot(aes(x = scenario_label, y = pct_extrapolation)) +
  geom_boxplot(fill = "lightyellow", outlier.shape = NA) +
  geom_jitter(aes(colour = gcm), width = 0.15, size = 2.5) +
  labs(
    title = "B) Extrapolation (MESS < 0) by scenario",
    x = NULL, y = "% cells with novel climate", colour = "GCM"
  ) +
  theme_minimal(base_size = 11)

p_box / p_extrap +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

ggsave(file.path(figures_dir, "10_fig3_gcm_uncertainty.png"),
       width = 10, height = 8, dpi = 300)
```


## 6. Priority surveillance map

Identify areas that are currently unsuitable but become suitable in ≥4 of 5
GCMs. These are high-confidence expansion zones worthy of targeted
surveillance.

```{r}
#| label: priority-surveillance

# Focus on the worst-case scenario for surveillance: SSP5-8.5, 2081-2100
# (most expansion, highest priority for monitoring)

scenario_paths <- read_csv(file.path(processed_v2, "future_scenario_paths.csv"),
                           show_col_types = FALSE)

# Find the 5 GCM predictions for SSP5-8.5, 2081-2100
target_scenarios <- scenario_paths %>%
  filter(ssp == "585", period == "2081-2100")

if (nrow(target_scenarios) >= 4) {
  # Stack binary predictions from all GCMs for this scenario
  gcm_binary_stack <- rast(lapply(target_scenarios$scenario_id, function(sid) {
    pred_file <- file.path(processed_v2, paste0("pred_future_", sid, ".tif"))
    pred <- rast(pred_file)
    pred >= primary_threshold
  }))

  # Count how many GCMs predict suitability at each cell
  n_gcms_suitable <- app(gcm_binary_stack, fun = "sum", na.rm = TRUE)
  names(n_gcms_suitable) <- "n_gcms"

  # Current binary
  current_suitable <- pred_current >= primary_threshold

  # Priority surveillance: currently unsuitable BUT suitable in ≥4 GCMs
  priority <- (current_suitable == 0) & (n_gcms_suitable >= 4)
  names(priority) <- "priority"

  # Calculate area
  cell_area_km2 <- prod(res(priority)) * 111^2 *
    cos(mean(ext(priority)[3:4]) * pi / 180)
  n_priority <- global(priority, sum, na.rm = TRUE)[[1]]
  area_priority <- n_priority * cell_area_km2

  sprintf("Priority surveillance cells: %.0f (%.0f km²)", n_priority, area_priority)

  # Map
  priority_df <- as.data.frame(priority, xy = TRUE) %>%
    filter(!is.na(priority), priority == 1)

  current_df <- as.data.frame(current_suitable, xy = TRUE) %>%
    filter(!is.na(suitability), suitability == 1)
}
```

```{r}
#| label: fig-priority-surveillance
#| fig-cap: "Figure 4. Priority surveillance areas — currently unsuitable but predicted suitable by ≥4 of 5 GCMs under SSP5-8.5 2081-2100."
#| fig-width: 10
#| fig-height: 8

if (exists("priority_df") && nrow(priority_df) > 0) {
  ggplot() +
    geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
    geom_raster(data = current_df, aes(x = x, y = y), fill = "#1f77b4", alpha = 0.4) +
    geom_raster(data = priority_df, aes(x = x, y = y), fill = "#ff7f0e") +
    coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
    labs(
      title = "Priority surveillance areas for H. longicornis",
      subtitle = sprintf(
        "Orange = new suitable habitat in ≥4/5 GCMs (SSP5-8.5, 2081–2100) — %.0f km²\nBlue = currently suitable habitat",
        area_priority
      )
    ) +
    theme_minimal(base_size = 12)

  ggsave(file.path(figures_dir, "10_fig4_priority_surveillance.png"),
         width = 10, height = 8, dpi = 300)
} else {
  message("No priority surveillance areas identified (or data not available)")
}
```

```{r}
#| label: fig-gcm-agreement
#| fig-cap: "Number of GCMs predicting suitability under SSP5-8.5 2081-2100."
#| fig-width: 10
#| fig-height: 8

if (exists("n_gcms_suitable")) {
  gcm_agree_df <- as.data.frame(n_gcms_suitable, xy = TRUE) %>%
    filter(!is.na(n_gcms), n_gcms > 0)

  ggplot() +
    geom_sf(data = world_map, fill = "grey95", colour = "grey70", linewidth = 0.2) +
    geom_raster(data = gcm_agree_df, aes(x = x, y = y, fill = factor(n_gcms))) +
    scale_fill_brewer(palette = "YlOrRd", name = "GCMs\nagreeing",
                      direction = 1) +
    coord_sf(xlim = c(110, 180), ylim = c(-48, -8)) +
    labs(
      title = "GCM agreement on habitat suitability (SSP5-8.5, 2081–2100)",
      subtitle = "Number of GCMs (out of 5) predicting suitable habitat"
    ) +
    theme_minimal(base_size = 12) +
    theme(legend.position = "right")

  ggsave(file.path(figures_dir, "10_fig5_gcm_agreement.png"),
         width = 10, height = 8, dpi = 300)
}
```


## 7. Summary tables

### Model performance comparison (v1 vs v2)

```{r}
#| label: model-comparison-table

comparison
```

### Range shift summary (all scenarios)

```{r}
#| label: range-shift-table

shift_table %>%
  mutate(across(where(is.numeric) & !contains("pct"), ~ round(., 0))) %>%
  select(scenario, current_suitable_km2, future_suitable_km2,
         gain_km2, loss_km2, net_change_km2, pct_change)
```

### Ensemble uncertainty

```{r}
#| label: uncertainty-table

uncertainty %>%
  select(ssp, period, median_pct_change, range_pct_change, median_extrapolation)
```

### Variable importance

```{r}
#| label: variable-importance-table

var_importance
```

### Biological validation

```{r}
#| label: validation-table

validation
```

### LORO cross-validation

```{r}
#| label: loro-table

loro_results
```


## 8. Save final outputs

```{r}
#| label: save-final

# Save priority surveillance raster if available
if (exists("priority")) {
  writeRaster(priority,
              file.path(processed_v2, "priority_surveillance.tif"),
              overwrite = TRUE)
  writeRaster(n_gcms_suitable,
              file.path(processed_v2, "gcm_agreement_ssp585_2081.tif"),
              overwrite = TRUE)
}

# Save a combined results summary
results_overview <- tibble(
  Section = c(
    rep("Model", 5),
    rep("Current prediction", 3),
    rep("Validation", 4),
    rep("Future (median across SSP×period)", nrow(shift_table))
  ),
  Metric = c(
    "Feature class", "Regularization", "AUC", "TSS", "Kappa",
    "Suitable area (km²)", "MESS extrapolation (%)", "Threshold (MaxSS)",
    "East coast suitability", "Desert suitability",
    "Schoener's D", "Warren's I",
    paste0("Net change (", shift_table$scenario, ")")
  ),
  Value = c(
    comparison$v2[comparison$Metric == "Feature class"],
    comparison$v2[comparison$Metric == "Regularization (β)"],
    as.character(metrics$Value[metrics$Metric == "AUC"]),
    as.character(metrics$Value[metrics$Metric == "TSS"]),
    as.character(metrics$Value[metrics$Metric == "Kappa"]),
    comparison$v2[comparison$Metric == "Background sampling"],
    sprintf("%.1f%%", 100 * global(mess_current < 0, sum, na.rm = TRUE)[[1]] /
              global(!is.na(mess_current), sum)[[1]]),
    sprintf("%.4f", primary_threshold),
    validation$Value[1],
    validation$Value[2],
    sprintf("%.3f", niche_overlap$Value[1]),
    sprintf("%.3f", niche_overlap$Value[2]),
    sprintf("%+.0f km² (%+.1f%%)", shift_table$net_change_km2, shift_table$pct_change)
  )
)

write_csv(results_overview, file.path(outputs_dir, "results_overview_v2.csv"))

sprintf("All visualization and results outputs saved")
```


## 9. Final interpretation notes

The following points should be included in any manuscript or report using
these results:

1. **The model represents a range shift, not simple expansion.** Under all
   future scenarios, some currently suitable habitat becomes unsuitable while
   new areas become suitable. The net change masks important spatial
   redistribution.

2. **Extrapolation uncertainty is present** — the MESS analysis shows that
   some proportion of Aus/NZ cells fall outside the training data's climate
   envelope. Predictions in these areas should be interpreted with caution.

3. **GCM choice matters.** The spread across 5 GCMs shows substantial
   uncertainty in the magnitude (and sometimes direction) of projected
   changes. Ensemble median is more robust than mean due to potential
   outlier GCMs.

4. **ENVIREM/CHELSA future variables were approximated** using delta-change
   methods rather than directly modeled by GCMs. This adds an additional
   layer of uncertainty, particularly for moisture-balance variables under
   scenarios with large precipitation changes.

5. **No dispersal constraints were applied.** The model assumes the species
   can colonize any newly suitable habitat immediately. In reality, expansion
   rates depend on livestock movement, wildlife corridors, and human-mediated
   transport. The suitable area maps should be interpreted as potential
   (not predicted) habitat.


## Session Information

```{r}
#| label: session-info
sessionInfo()
```
